# Time series forecasting with machine learning in R

```{r echo=FALSE}
# This removes all items in environment.
rm(list = ls())
```

As in the previous chapter, the problem is to forecast one monthly time series:
*Beer, Wine, and Distilled Alcoholic Beverages Sales*, following the original
Matt Dancho example. The data comes from FRED, the Federal Reserve Economic
Data database. The series belongs to the non-durable goods category; it
includes U.S. merchant wholesalers, excluding manufacturers' sales branches and
offices. The observations run from 2010-01-01 to 2022-10-31. The goal is to
estimate models with information available before 2022 and evaluate their
forecasts over the 10 available months of 2022.

For the full database details see:
https://fred.stlouisfed.org/series/S4248SM144NCEN

First load the R packages.

```{r}
# Load libraries.
library(fpp3)
library(timetk)     # Toolkit for working with time series in R.
library(tidyquant)  # Loads tidyverse, financial packages, and data tools.
library(dplyr)      # Database manipulation.
library(ggplot2)    # Plots.
library(tibble)     # Tables.
library(kableExtra) # Tables.
library(knitr)
library(sweep)      # Broom-style tidiers for the forecast package.
library(forecast)   # Forecasting models and predictions package.
library(seasonal)
library(lubridate)
library(rpart)      # Lightweight regression trees.
library(nnet)       # Lightweight neural networks.
```

```{r}
# This comes from 01-forecasting-with-fpp3.qmd.
mape_updated <- readRDS("mape_updated.rds")
```

We can conveniently download the data directly from the FRED API in one line
of code.

```{r get-data-ml}
# Beer, Wine, Distilled Alcoholic Beverages, in Millions USD.
beer <- tq_get("S4248SM144NCEN", get = "economic.data",
               from = "2010-01-01", to = "2022-10-31")
```

The first rows show the structure of the data set. By default the value column
is named *price*, although these are sales figures in monetary terms. According
to the main FRED reference, these are in millions of dollars and are reported
before seasonal adjustment.

```{r see-data-ml}
head(beer)
```

We rename the value column so that the code matches the economic meaning of the
variable.

```{r rename-sales-ml}
beer <- beer %>%
  rename(sales = price)

tail(beer)
```

The column name now identifies the variable as sales.

## Supervised forecasting workflow

In this chapter, machine learning is used as a forecasting tool. The goal is
narrowly defined: forecast a monthly time series by turning calendar and trend
information into predictors. In time series forecasting, the supervised
learning idea is to transform an ordered sequence into a modeling table: build
a table of predictors called $X$, keep the variable we want to forecast as
$y$, estimate a model on the training data, tune it with validation data, and
evaluate it once on the test data.

This chapter uses the same forecasting problem as before with an explicit
supervised forecasting workflow. The chapter writes the model comparison directly:
we create the predictors, choose the candidate model families, tune the few
hyperparameters by hand, and evaluate the final models on a holdout period. The
models are smaller, faster to render in Quarto, easier to explain, and easier
to debug. The workflow is also close to what students need to understand:

1. create time-based features;
2. split the data without shuffling the time order;
3. fit several explicit models;
4. choose hyperparameters with the validation set;
5. evaluate the chosen models on a test set held apart from fitting and tuning.

We will compare three transparent machine-learning forecasting alternatives:

- a multivariate linear regression;
- a regression tree;
- a small neural network.

The first model is highly interpretable. The tree is nonlinear and can capture
simple threshold behavior. The neural network is also nonlinear. We keep it
small so the model remains pedagogical and reproducible.

Mathematically, this chapter changes the forecasting problem from a univariate
time-series model into a supervised forecasting regression problem. The first
chapter used models built directly from the history of the series. This chapter
turns each month into one row of a modeling table:

| Row element | Meaning |
|---|---|
| `sales` | The value we want to forecast. |
| `index.num` | A numeric time trend. |
| `year` | A coarse trend variable and split variable. |
| `month.lbl` | Monthly seasonality. |

Then each supervised forecasting model estimates a rule

$$
\text{sales} = f(\text{trend and month variables}) + \text{error}.
$$

The three alternatives below differ in how flexible we allow the rule to be.
Linear regression uses an additive straight-line rule. The regression tree uses
if-then splits. The neural network uses a small nonlinear rule. The transparency
principle is simple: we can see the columns, we can see the candidate model
families, and the test set is reserved for the final evaluation.

The data split has three different jobs:

| Sample | Dates | Job in the workflow |
|---|---|---|
| Training | 2010-2020 | Estimate model parameters. |
| Validation | 2021 | Choose hyperparameters such as tree complexity or neural-network size. |
| Test | January-October 2022 | Evaluate the selected models once. |

## The data

We need an additional data set. The *validation* data set is used for model
selection: it helps us choose hyperparameters before touching the final test
period. The new data sets are the following.

```{r}
#| label: fig-bstvats
#| fig-cap: "Beer sales: train, validation and test sets."
beer %>%
  ggplot(aes(date, sales)) +
  annotate("text", x = ymd("2013-01-01"), y = 14000,
           color = "black", label = "Train region") +
  geom_rect(xmin = as.numeric(ymd("2021-01-01")),
            xmax = as.numeric(ymd("2021-12-31")), ymin = 0, ymax = 20000,
            alpha = 0.01, fill = "purple") +
  annotate("text", x = ymd("2021-04-01"), y = 7000,
           color = "black", label = "Validation\nregion") +
  geom_rect(xmin = as.numeric(ymd("2022-01-01")),
            xmax = as.numeric(ymd("2022-09-30")), ymin = 0, ymax = 20000,
            alpha = 0.02, fill = "pink") +
  annotate("text", x = ymd("2022-06-01"), y = 9000,
           color = "black", label = "Test\nregion") +
  geom_line(col = "black") +
  geom_point(col = "black", alpha = 0.5, size = 2) +
  theme_tq() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(subtitle =
  "Train (2010 - 2020), validation (2021), and test set (Jan 2022 to Oct 2022)",
  x = "Date", y = "Sales",
       caption = "The test region is hidden from the models during fitting;
       it is shown here to evaluate the 10-month ahead forecast.") +
  scale_y_continuous(labels = scales::dollar)
```

And the corresponding zoom version.

```{r}
#| label: fig-zbstvats
#| fig-cap: "Zoom. Beer sales: train, validation and test sets."
beer %>%
  filter(date > as.Date("2020-01-01")) %>%
  ggplot(aes(date, sales)) +
  annotate("text", x = ymd("2020-08-01"), y = 14000,
           color = "black", label = "Train region") +
  geom_rect(xmin = as.numeric(ymd("2021-01-01")),
            xmax = as.numeric(ymd("2021-12-31")), ymin = 0, ymax = 20000,
            alpha = 0.01, fill = "purple") +
  annotate("text", x = ymd("2021-07-01"), y = 14000,
           color = "black", label = "Validation region") +
  geom_rect(xmin = as.numeric(ymd("2022-01-01")),
            xmax = as.numeric(ymd("2022-09-30")), ymin = 0, ymax = 20000,
            alpha = 0.02, fill = "pink") +
  annotate("text", x = ymd("2022-05-01"), y = 14000,
           color = "black", label = "Test region") +
  geom_line(col = "black") +
  geom_point(col = "black", alpha = 0.5, size = 5) +
  theme_tq() +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  labs(subtitle =
  "Train (2010 - 2020), validation (2021), and test set (Jan 2022 to Oct 2022)",
  x = "Date", y = "Sales",
       caption = "The test region is hidden from the models during fitting;
       it is shown here to evaluate the 10-month ahead forecast.") +
  scale_y_continuous(labels = scales::dollar)
```

The `tk_augment_timeseries_signature()` function expands the timestamp
information column-wise into a forecasting feature set. The original data
set has a date, a symbol, and the sales variable. The augmented data includes
calendar information such as year, month, quarter, semester and other time
features. Only some of them will be useful for the final models. The full list
still helps us inspect the forecasting problem as a supervised forecasting
problem.

```{r data-in-a-forecasting-feature-format}
beer_aug <- beer %>%
  tk_augment_timeseries_signature()

tail(beer_aug)
```

The variable names in `beer_aug` are:

```{r}
colnames(beer_aug)
```

The full signature is useful for inspection. For this monthly data set, it is
broader than necessary. Many columns describe hours, seconds, weekdays, or
day-of-month details. They are meaningful for hourly or daily data. In this
monthly problem, they mostly encode calendar artifacts. We keep a smaller set:

- `sales`, the target variable;
- `index.num`, a numeric time index for the trend;
- `year`, a coarse trend variable that also makes the split readable;
- `month.lbl`, a factor for monthly seasonality.

This modeling decision keeps the target variable intact and makes the
supervised forecasting matrix easier to explain. It also reduces dependence on
irrelevant timestamp fragments.

The retained features define the design matrix used by the models. Because
`month.lbl` is a categorical variable, R expands it internally into 0/1 dummy
variables. For example, define

$$
D_{t,\text{Feb}} =
\begin{cases}
1, & \text{month } t \text{ is February},\\
0, & \text{otherwise.}
\end{cases}
$$

The same idea is repeated for March through December. Conceptually, the row for
month $t$ is

$$
x_t =
\left[
1,\,
\text{index.num}_t,\,
\text{year}_t,\,
D_{t,\text{Feb}},\ldots,
D_{t,\text{Dec}}
\right].
$$

January is absorbed into the intercept by default. The important forecasting
point is that every predictor in $x_t$ is known before the sales value $y_t$ is
observed. Future dates, years, and months are known in advance, so these
predictors keep the future sales value unavailable to the model.

```{r feature-selection-rationale}
selected_ml_columns <- c("sales", "index.num", "year", "month.lbl")

feature_decisions <- tibble(variable = colnames(beer_aug)) %>%
  mutate(
    decision = case_when(
      variable == "sales" ~ "Target",
      variable %in% selected_ml_columns ~ "Keep",
      TRUE ~ "Drop"
    ),
    reason = case_when(
      variable == "sales" ~ "Value we want to forecast.",
      variable == "index.num" ~ "Numeric trend over time.",
      variable == "year" ~ "Coarse trend and readable split variable.",
      variable == "month.lbl" ~ "Monthly seasonal factor.",
      variable == "date" ~ "Raw Date object; useful for plots; excluded as a direct predictor here.",
      variable == "symbol" ~ "Constant because there is only one series.",
      variable %in% c("hour", "minute", "second", "hour12", "am.pm") ~
        "Intraday information; irrelevant for monthly observations.",
      variable %in% c("day", "wday", "wday.xts", "wday.lbl", "mday",
                      "qday", "yday", "mweek", "week", "week.iso",
                      "week2", "week3", "week4", "mday7") ~
        "Daily or weekly calendar artifact for monthly data.",
      TRUE ~ "Redundant with the retained trend or monthly seasonality variables."
    )
  )

kable(feature_decisions,
      caption = "Feature-selection decision for the time-series signature.",
      row.names = FALSE) %>%
  kable_styling(latex_options = "HOLD_position")
```

We now build the actual modeling table from the selected variables and convert
ordered factors into plain factors. The target variable `sales` remains
unchanged; the code only prepares the predictor matrix.

```{r data-cleaning}
beer_clean <- beer_aug %>%
  select(all_of(selected_ml_columns)) %>%
  mutate_if(is.ordered, ~ as.character(.) %>% as.factor)

head(beer_clean)
```

We split the database into training, validation and test sets following the
time ranges in the visualization above. The test set is the most recent data.
It remains unavailable during model fitting and tuning.

The split can be read directly from the `year` column:

| Sample | Rule in the code | Use |
|---|---|---|
| Training | `year < 2021` | Fit model parameters. |
| Validation | `year == 2021` | Choose hyperparameters. |
| Test | `year == 2022` | Evaluate the chosen models. |

The validation set is used to choose hyperparameters. The test set is used only
after the choices have been made.

```{r split-ml-data}
train_tbl <- beer_clean %>% filter(year < 2021)
valid_tbl <- beer_clean %>% filter(year == 2021)
test_tbl  <- beer_clean %>% filter(year == 2022)

test_tbl$sales
```

Our goal is to forecast the first 10 months of 2022.

## Validation workflow

The following support functions keep the comparison compact. They calculate the
same error measures for every model: mean error, root mean squared error, mean
absolute error, mean absolute percentage error, and mean percentage error.

For a validation or test sample of size $n$, with predictions $\hat{y}_t$, the
main metrics are:

$$
\text{ME} = \frac{1}{n}\sum_{t=1}^{n}(y_t-\hat{y}_t),
$$

$$
\text{RMSE} =
\sqrt{\frac{1}{n}\sum_{t=1}^{n}(y_t-\hat{y}_t)^2},
$$

$$
\text{MAE} =
\frac{1}{n}\sum_{t=1}^{n}|y_t-\hat{y}_t|,
$$

$$
\text{MAPE} =
\frac{100}{n}\sum_{t=1}^{n}
\left|
\frac{y_t-\hat{y}_t}{y_t}
\right|.
$$

The code below implements these formulas directly. This keeps the workflow
auditable because the evaluation formulas are visible in the chapter.

```{r metric-functions}
metric_tbl <- function(actual, pred) {
  residuals <- actual - pred
  pct_error <- residuals / actual

  tibble(
    me = mean(residuals, na.rm = TRUE),
    rmse = sqrt(mean(residuals^2, na.rm = TRUE)),
    mae = mean(abs(residuals), na.rm = TRUE),
    mape = 100 * mean(abs(pct_error), na.rm = TRUE),
    mpe = 100 * mean(pct_error, na.rm = TRUE)
  )
}

validation_row <- function(model, family, setting, actual, pred) {
  bind_cols(
    tibble(model = model, family = family, setting = setting),
    metric_tbl(actual, pred)
  )
}

forecast_error_tbl <- function(model, pred) {
  beer %>%
    filter(lubridate::year(date) == 2022) %>%
    transmute(
      model = model,
      date = date,
      actual = sales,
      pred = as.numeric(pred),
      error = actual - pred,
      error_pct = error / actual
    )
}
```

The linear regression has no hyperparameters. It is our transparent baseline.
If this model works well, that is useful information: sometimes a simple model
is enough when trend and seasonality are strong.

The fitted linear regression equation is

$$
y_t =
\beta_0
+ \beta_1\,\text{index.num}_t
+ \beta_2\,\text{year}_t
+ \sum_{j=2}^{12}\gamma_j
D_{t,j}
+ \varepsilon_t.
$$

Here, $D_{t,j}$ is a monthly dummy variable: it equals 1 when observation $t$
belongs to month $j$, and 0 otherwise. January is the baseline month because it
is absorbed by the intercept. The coefficients have a direct interpretation:
the trend variables explain long-run movement, and the monthly dummy variables
explain seasonality relative to January. The code `lm(sales ~ .)` means:
regress `sales` on all remaining columns in the modeling table.

```{r validate-lm}
ml_formula <- sales ~ .

fit_lm_valid <- lm(ml_formula, data = train_tbl)
pred_lm_valid <- as.numeric(predict(fit_lm_valid, newdata = valid_tbl))

lm_validation <- validation_row(
  model = "Linear regression",
  family = "Linear regression",
  setting = "selected calendar features",
  actual = valid_tbl$sales,
  pred = pred_lm_valid
)
```

The regression tree has a complexity parameter, `cp`. A smaller value allows a
larger tree; a larger value prunes the tree more aggressively. We estimate a
small grid and choose the value that minimizes validation RMSE.

A regression tree asks a sequence of yes/no questions about the predictors.
For example: is the month December? is the time index above a certain value?
After following those questions to a final group, the forecast is the average
sales value among the training observations in that group:

$$
\text{tree forecast}
=
\text{average sales in the final group}.
$$

The `cp` parameter controls the cost of adding more splits. A useful way to
read it is as a penalty on tree complexity: larger `cp` values prune more
splits, and smaller `cp` values allow more splits.

Small `cp` values allow more regions and can overfit. Large `cp` values produce
simpler trees and can underfit. We choose `cp` using the validation set. The
test set remains reserved for the final evaluation.

```{r validate-tree}
tree_grid <- tibble(cp = c(0.001, 0.005, 0.01, 0.02, 0.05))

tree_validation <- bind_rows(lapply(seq_len(nrow(tree_grid)), function(i) {
  fit_tree <- rpart::rpart(
    ml_formula,
    data = train_tbl,
    method = "anova",
    control = rpart::rpart.control(
      cp = tree_grid$cp[i],
      minsplit = 8,
      xval = 0
    )
  )

  pred_tree <- as.numeric(predict(fit_tree, newdata = valid_tbl))

  validation_row(
    model = "Regression tree",
    family = "Regression tree",
    setting = paste0("cp = ", tree_grid$cp[i]),
    actual = valid_tbl$sales,
    pred = pred_tree
  )
}))

best_tree <- tree_validation %>%
  arrange(rmse) %>%
  slice(1)
```

The neural network also receives a small manual search. We tune two
hyperparameters: `size`, the number of hidden units, and `decay`, a penalty
that discourages overly large weights. Because the variables have different
scales, we standardize the model matrix before fitting.

The neural network uses the same predictors as the linear regression and tree.
The difference is the way it combines them. It first standardizes the inputs,
then creates a small number of hidden units, and finally combines those hidden
units into one forecast. The `size` hyperparameter controls how many hidden
units are used. The `decay` hyperparameter discourages very large weights, so
the fitted rule stays smoother.

Its inputs remain the same explicit features used by the linear regression and
the tree. The extra flexibility comes from the hidden layer and the nonlinear
activation function.

The support functions below do five mechanical tasks:

| Step | What the code does | Why it is important |
|---|---|---|
| 1 | Build the model matrix from the formula. | Convert factors such as month into numeric columns. |
| 2 | Standardize the predictors. | Put all inputs on comparable scales. |
| 3 | Standardize the target variable. | Help the neural network optimize more smoothly. |
| 4 | Fit the network with `nnet()`. | Estimate the weights in the hidden and output layers. |
| 5 | Undo the scaling during prediction. | Return forecasts in the original sales units. |

```{r nnet-functions}
fit_nnet_regression <- function(train_data, size, decay, seed = 13) {
  x <- model.matrix(ml_formula, data = train_data)[, -1, drop = FALSE]

  center <- apply(x, 2, mean)
  scale <- apply(x, 2, sd)
  scale[is.na(scale) | scale == 0] <- 1

  x_scaled <- sweep(sweep(x, 2, center, "-"), 2, scale, "/")

  y_center <- mean(train_data$sales)
  y_scale <- sd(train_data$sales)
  if (is.na(y_scale) || y_scale == 0) y_scale <- 1

  set.seed(seed)
  fit <- nnet::nnet(
    x = x_scaled,
    y = (train_data$sales - y_center) / y_scale,
    size = size,
    decay = decay,
    linout = TRUE,
    trace = FALSE,
    maxit = 1000
  )

  list(
    fit = fit,
    columns = colnames(x),
    center = center,
    scale = scale,
    y_center = y_center,
    y_scale = y_scale
  )
}

predict_nnet_regression <- function(object, new_data) {
  x <- model.matrix(ml_formula, data = new_data)[, -1, drop = FALSE]

  missing_cols <- setdiff(object$columns, colnames(x))
  if (length(missing_cols) > 0) {
    missing_matrix <- matrix(0, nrow = nrow(x), ncol = length(missing_cols))
    colnames(missing_matrix) <- missing_cols
    x <- cbind(x, missing_matrix)
  }

  x <- x[, object$columns, drop = FALSE]
  x_scaled <- sweep(sweep(x, 2, object$center, "-"), 2, object$scale, "/")

  as.numeric(predict(object$fit, x_scaled)) * object$y_scale + object$y_center
}
```

```{r validate-nnet}
nnet_grid <- expand.grid(
  size = c(1, 3, 5),
  decay = c(0, 0.001, 0.01)
)

nnet_validation <- bind_rows(lapply(seq_len(nrow(nnet_grid)), function(i) {
  fit_nnet <- fit_nnet_regression(
    train_data = train_tbl,
    size = nnet_grid$size[i],
    decay = nnet_grid$decay[i],
    seed = 13 + i
  )

  pred_nnet <- predict_nnet_regression(fit_nnet, valid_tbl)

  validation_row(
    model = "Neural network",
    family = "Neural network",
    setting = paste0("size = ", nnet_grid$size[i],
                     ", decay = ", nnet_grid$decay[i]),
    actual = valid_tbl$sales,
    pred = pred_nnet
  )
}))

best_nnet <- nnet_validation %>%
  arrange(rmse) %>%
  slice(1)
```

The validation leaderboard summarizes the model selection step. The final
forecasting performance is evaluated later with the test set.

For each model family, we choose the setting with the smallest validation RMSE.
For the tree, the setting is `cp`. For the neural network, the setting is the
pair (`size`, `decay`). Linear regression has no tuning grid, so it enters the
leaderboard once.

```{r validation-leaderboard}
validation_leaderboard <- bind_rows(
  lm_validation,
  tree_validation,
  nnet_validation
) %>%
  arrange(rmse)

kable(validation_leaderboard,
      caption = "Validation leaderboard for machine-learning forecasting models.",
      digits = 2) %>%
  kable_styling(latex_options = "HOLD_position")
```

## Test forecasts

Now we refit the chosen model specifications using the training and validation
data together, from 2010 through 2021. Then we forecast the 2022 test set. This
is a standard practice: validation is used to make modeling decisions, and
after that decision is made we can use all pre-test information to estimate
the final model.

The final fitted function uses all observations from 2010 through 2021 and the
hyperparameters selected in validation. The test forecasts are obtained by
applying that final fitted rule to the 2022 rows.

This ordering prevents methodological leakage: the test sales values never
enter coefficient estimation, `cp` selection, neural-network size selection, or
neural-network decay selection.

```{r fit-final-ml-models}
train_valid_tbl <- beer_clean %>%
  filter(year < 2022)

fit_lm_final <- lm(ml_formula, data = train_valid_tbl)
pred_lm_test <- as.numeric(predict(fit_lm_final, newdata = test_tbl))

fit_tree_final <- rpart::rpart(
  ml_formula,
  data = train_valid_tbl,
  method = "anova",
  control = rpart::rpart.control(
    cp = best_tree$setting %>%
      sub("cp = ", "", .) %>%
      as.numeric(),
    minsplit = 8,
    xval = 0
  )
)
pred_tree_test <- as.numeric(predict(fit_tree_final, newdata = test_tbl))

best_nnet_size <- sub("size = ([0-9]+),.*", "\\1", best_nnet$setting) %>%
  as.numeric()
best_nnet_decay <- sub(".*decay = ([0-9.]+)", "\\1", best_nnet$setting) %>%
  as.numeric()

fit_nnet_final <- fit_nnet_regression(
  train_data = train_valid_tbl,
  size = best_nnet_size,
  decay = best_nnet_decay,
  seed = 100
)
pred_nnet_test <- predict_nnet_regression(fit_nnet_final, test_tbl)
```

The table below shows the 10-month forecast errors.

```{r ml-error-table}
ml_errors <- bind_rows(
  forecast_error_tbl("ML forecast: linear regression", pred_lm_test),
  forecast_error_tbl("ML forecast: regression tree", pred_tree_test),
  forecast_error_tbl("ML forecast: neural network", pred_nnet_test)
)

kable(ml_errors,
      caption = "Detailed test performance for machine-learning forecasting models.",
      digits = 3,
      row.names = FALSE) %>%
  kable_styling(latex_options = "HOLD_position")
```

The same information is clearer in a plot. The black line is the actual series.
The colored lines are the machine-learning forecasts for the 2022 test period.

```{r}
#| label: fig-transparent-ml-forecasts
#| fig-cap: "Forecasts from machine-learning forecasting models."
beer %>%
  filter(date > as.Date("2021-01-01")) %>%
  ggplot(aes(x = yearmonth(date), y = sales)) +
  geom_line(color = "black", linewidth = 0.5) +
  geom_point(color = "black", size = 4, alpha = 0.5) +
  geom_line(aes(y = pred, color = model), linewidth = 0.7,
            data = ml_errors) +
  geom_point(aes(y = pred, color = model), size = 3, alpha = 0.8,
             data = ml_errors) +
  geom_vline(xintercept = as.numeric(as.Date("2021-12-01")), linetype = 2) +
  theme_tq() +
  labs(x = "Date", y = "Sales", color = "Model") +
  scale_y_continuous(labels = scales::dollar)
```

The overall test summary is:

```{r ml-summary}
ml_summary <- ml_errors %>%
  group_by(model) %>%
  summarise(
    me = mean(error),
    rmse = sqrt(mean(error^2)),
    mae = mean(abs(error)),
    mape = 100 * mean(abs(error_pct)),
    mpe = 100 * mean(error_pct),
    .groups = "drop"
  ) %>%
  arrange(mape)

kable(ml_summary,
      caption = "Test summary for machine-learning forecasting models.",
      digits = 2) %>%
  kable_styling(latex_options = "HOLD_position")
```

A model can rank first by MAPE and rank differently by another metric. For that
reason, it is useful to inspect multiple errors. RMSE penalizes large mistakes
more heavily, while MAPE is easy to read as an average percentage error.

The table compares fitted forecasting functions under a common experimental
design. Every model saw the same training data, the same validation data, the
same predictors, and the same test months. A lower MAPE means that this
particular fitted function was closer to the realized 2022 sales over this
holdout period.

In the current render, the machine-learning neural network forecast is the best
of the three supervised forecasting models, with MAPE around 4.91. The
regression tree follows with MAPE around 5.16, and the linear regression has
MAPE around 5.81. The differences are modest, so the interpretation should
focus on the pattern: calendar and trend predictors carry useful information,
and a small nonlinear model extracts a little more from those predictors in
this test window.

We can also show the best individual point forecast for each model.

```{r point-forecast-ml}
point_forecast_ml <- ml_errors %>%
  group_by(model) %>%
  slice_min(abs(error_pct), n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(model, date, actual, pred, error, error_pct)

kable(point_forecast_ml,
      caption = "Best point forecast for each machine-learning forecasting model.",
      digits = 4,
      row.names = FALSE) %>%
  kable_styling(latex_options = "HOLD_position")
```

Choosing a model according to one specific point forecast would be fragile.
Even so, it is useful to see where each model was most accurate.

```{r update-mape-ml}
mape_updated_ml <- mape_updated %>%
  add_row(.model = "ML forecast: linear regression",
          MAPE = ml_summary$mape[ml_summary$model == "ML forecast: linear regression"]) %>%
  add_row(.model = "ML forecast: regression tree",
          MAPE = ml_summary$mape[ml_summary$model == "ML forecast: regression tree"]) %>%
  add_row(.model = "ML forecast: neural network",
          MAPE = ml_summary$mape[ml_summary$model == "ML forecast: neural network"]) %>%
  arrange(MAPE)

mape_updated_ml
```

## ARIMA comparison

As in the original Matt Dancho example, we include a different ARIMA workflow
using the `forecast` package and the `auto.arima()` function. This benchmark
belongs to the classical time-series family and answers a different modeling
question. The machine-learning forecasting models above estimate
$\text{sales}=f(\text{trend and month variables})+\text{error}$ using trend
and calendar predictors. ARIMA models the dependence of the series on its own
past.

The `auto.arima()` function automates the selection of the ARIMA orders. The
automation is a statistical search over candidate values of
$(p,d,q)(P,D,Q)_{12}$, followed by parameter estimation and information-criterion
comparison.

The order notation is read as follows:

| Order | Meaning |
|---|---|
| $p$ | Number of recent lagged values used by the autoregressive part. |
| $d$ | Number of ordinary differences. |
| $q$ | Number of recent forecast errors used by the moving-average part. |
| $P$ | Number of seasonal lagged values. |
| $D$ | Number of seasonal differences. |
| $Q$ | Number of seasonal forecast errors. |
| 12 | Monthly seasonal cycle. |

**Prepare the data.**

The `tk_ts()` function coerces time series objects and tibbles with date or
date-time columns to `ts`.

```{r}
beer_sales_ts <- tk_ts(
  beer %>% filter(date < as.Date("2022-01-01")),
  start = 2010,
  freq = 12
)
beer_sales_ts
```

Just verify that the `tk_ts()` function worked.

```{r}
has_timetk_idx(beer_sales_ts)
```

**Estimation.**

We can use the `auto.arima()` function from the `forecast` package to model the
time series. The function tests a set of specifications and selects the one
that performs best according to its internal statistical criterion.

```{r}
set.seed(13)
fit_arima <- auto.arima(beer_sales_ts)
fit_arima
```

The printed output is the selected mathematical specification. After this line,
we are no longer speaking abstractly about `auto.arima()`; we are speaking about
the particular ARIMA equation selected for this data set and this training
window.

The `sw_tidy()` function returns the model coefficients in a tibble.

```{r}
sw_tidy(fit_arima)
```

The `sw_glance()` function returns training-set accuracy measures.

```{r}
sw_glance(fit_arima) %>%
  glimpse()
```

**Forecast.**

The `sw_augment()` function helps with model evaluation. We get the `.actual`,
`.fitted` and `.resid` columns, which are useful in evaluating the model
against the training data.

```{r}
sw_augment(fit_arima, timetk_idx = TRUE)
```

We can visualize the residual diagnostics for the training data.

```{r}
#| label: fig-fard
#| fig-cap: "forecast::auto.arima residual diagnosis."
sw_augment(fit_arima, timetk_idx = TRUE) %>%
  ggplot(aes(x = index, y = .resid)) +
  geom_hline(yintercept = 0, color = "red", linewidth = 1) +
  geom_point(size = 4, alpha = 0.5) +
  labs(x = "Date", y = "Residuals") +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  theme_tq() +
  scale_y_continuous(labels = scales::dollar)
```

The residuals are

$$
\hat{\varepsilon}_t = y_t - \hat{y}_{t\mid t-1}.
$$

If the ARIMA model has captured the predictable autocorrelation, these
residuals should have no obvious remaining time pattern. This is the same audit
logic used in the `fpp3` chapter.

Make a forecast using the `forecast()` function.

```{r}
fcast_arima <- forecast(fit_arima, h = 10)
fcast_arima
```

We use `sw_sweep()` to tidy the forecast output.

```{r}
fcast_tbl <- sw_sweep(fcast_arima, timetk_idx = TRUE)
tail(fcast_tbl, 10)
```

We can investigate the error on our test set.

```{r}
actuals_tbl <- beer %>%
  filter(lubridate::year(date) == 2022) %>%
  select(date, sales)

error_tbl_arima <- left_join(actuals_tbl, fcast_tbl,
                             by = c("date" = "index")) %>%
  rename(actual = sales.x, pred = sales.y) %>%
  select(date, actual, pred) %>%
  mutate(error = round((actual - pred), 2),
         error_pct = round((error / actual), 4))

error_tbl_arima
```

And we can calculate the same error metrics.

```{r}
metric_tbl(error_tbl_arima$actual, error_tbl_arima$pred) %>%
  glimpse()
```

Notice that we have the entire forecast in a tibble. We can now more easily
visualize the forecast.

```{r}
#| label: fig-fa
#| fig-cap: "Forecast. forecast::auto.arima."
fcast_tbl %>%
  filter(index > as.Date("2021-01-01")) %>%
  ggplot(aes(x = index, y = sales, color = key)) +
  geom_ribbon(aes(ymin = lo.95, ymax = hi.95),
              fill = "#D5DBFF", color = NA, linewidth = 0) +
  geom_ribbon(aes(ymin = lo.80, ymax = hi.80, fill = key),
              fill = "#596DD5", color = NA, linewidth = 0, alpha = 0.8) +
  geom_line() +
  geom_point(size = 4, alpha = 0.5) +
  geom_line(aes(x = date, y = sales), color = "black",
            data = actuals_tbl) +
  geom_point(aes(x = date, y = sales), color = "black", size = 4,
             alpha = 0.5, data = actuals_tbl) +
  geom_vline(xintercept = as.numeric(as.Date("2021-12-01")),
             linetype = 2) +
  labs(x = "Date", y = "Sales",
       caption = paste("MAPE =",
                       round(100 * mean(abs(error_tbl_arima$error_pct)), 5))) +
  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
  scale_color_tq() +
  scale_fill_tq() +
  theme_tq() +
  scale_y_continuous(labels = scales::dollar)
```

The resulting forecast serves as a benchmark.

In this render, `forecast::auto.arima` has MAPE around 3.69. This makes it a
strong benchmark: it beats the supervised machine-learning forecasts in this
specific holdout period and lands close to the best models from the first
chapter. That result is pedagogically useful because it shows that classical
time-series structure can remain highly competitive when the series has strong
trend, seasonality and autocorrelation.

## Summary of all results

An interesting question is: *What happens to the accuracy when you average the
predictions of different methods?* This question makes sense because model
choice is often uncertain in forecasting. Taking the average can reduce extreme
results, although it can also be harder to interpret because the final forecast
comes from several techniques.

If there are $M$ forecast models, the average forecast is

$$
\hat{y}^{\text{avg}}_t =
\frac{1}{M}\sum_{m=1}^{M}\hat{y}^{(m)}_t.
$$

This ensemble rule is applied after the individual forecasts already exist. Its
appeal is that different models can make different errors, so averaging may
reduce model-specific noise. Its cost is interpretability: the final number no
longer comes from one clear structural assumption.

In order to calculate the average forecast we need to load the forecasts from
the previous section. In particular, we include the seasonal naive,
NNAR(15,1,8)[12], ETS(M,A,M), and ARIMA(4,1,1)(0,1,1)[12]. We exclude the
naive, drift and mean forecasts from this average.

```{r}
fpp3_fc <- readRDS("fpp3_fc.rds")
```

We add the machine-learning forecasts and the `forecast::auto.arima`
forecast into the `fpp3` forecast table.

```{r}
ml_fc <- ml_errors %>%
  transmute(.model = model, date = yearmonth(date), .mean = pred) %>%
  as_tsibble(key = .model, index = date)

pred_arima_fc <- error_tbl_arima %>%
  transmute(.model = "forecast::auto.arima",
            date = yearmonth(date),
            .mean = pred) %>%
  as_tsibble(key = .model, index = date)

all_fc <- as_tsibble(fpp3_fc) %>%
  select(.model, date, .mean) %>%
  bind_rows(ml_fc, pred_arima_fc)

average <- all_fc %>%
  index_by(date) %>%
  summarise(av_fc = mean(.mean, na.rm = TRUE)) %>%
  bind_cols(sales = actuals_tbl$sales) %>%
  mutate(error = round((sales - av_fc), 2),
         error_pct = round((error / sales), 4))

average
```

Finally, the average forecast.

```{r}
#| label: fig-fav
#| fig-cap: "Forecast. Average."
beer %>%
  filter(date > as.Date("2021-01-01")) %>%
  ggplot(aes(x = yearmonth(date), y = sales)) +
  geom_line(color = "black", linewidth = 0.5) +
  geom_point(color = "black", size = 4, alpha = 0.5) +
  geom_line(aes(y = av_fc), linewidth = 0.5,
            color = "red", data = average) +
  geom_point(aes(y = av_fc), size = 4, alpha = 0.5,
             color = "red", data = average) +
  geom_vline(xintercept = as.numeric(as.Date("2021-12-01")), linetype = 2) +
  theme_tq() +
  labs(x = "Date", y = "Sales",
       caption = paste("MAPE =",
                       round(100 * mean(abs(average$error_pct)), 5))) +
  scale_y_continuous(labels = scales::dollar)
```

The final MAPE summary is updated with the ARIMA benchmark and the average
forecast.

```{r}
mape_updated_all <- mape_updated_ml %>%
  add_row(.model = "forecast::auto.arima",
          MAPE = 100 * mean(abs(error_tbl_arima$error_pct))) %>%
  add_row(.model = "Average forecast",
          MAPE = 100 * mean(abs(average$error_pct))) %>%
  arrange(MAPE)

mape_updated_all
```

The ranking is local to this data set and this holdout period. The main
lesson is that a transparent workflow lets us see where the forecast comes
from, how the model was selected, how long it takes to render, and how it
performs when confronted with truly unseen data.

The average forecast is the best row in this render, with MAPE around 2.37.
That result has a clear interpretation: the individual models make different
errors, and averaging reduces part of that model-specific noise. The cost is
interpretability because the final number comes from several forecasting
rules. For teaching purposes, this is a valuable ending point: students can
see that the best empirical forecast may come from combining models, while the
individual models remain necessary for understanding what information the
forecast is using.

## References

The main web references of this document are:

- [Time Series Machine Learning with timetk.](https://www.business-science.io/code-tools/2017/10/24/demo_week_timetk.html)
- [Tidy Forecasting with sweep.](https://www.business-science.io/code-tools/2017/10/25/demo_week_sweep.html)
- [Forecasting: Principles and Practice.](https://otexts.com/fpp3/)
