# GARCH volatility forecasting in R

```{r}
# This removes all items in environment.
rm(list = ls())

library(tidyquant)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(knitr)
library(kableExtra)
library(scales)
```

## Volatility as a forecasting problem

The previous chapters forecast the level of a monthly sales series. Here the
forecast target changes. We work with daily financial returns and forecast
**conditional variance**, the variance expected for the next period using the
information available today. The implementation is adapted from the original
pedagogical GARCH material in @lozano2025garchvideo and rewritten here as a
Quarto book chapter focused on applied forecasting.

In financial time series, large movements tend to cluster. Calm periods can
last for many days, and turbulent periods can also persist. A single
unconditional standard deviation summarizes the full sample with one number.
A GARCH model produces a time series of conditional volatility, which is useful
precisely because risk changes through time. This risk-management motivation is
standard in financial engineering and quantitative risk management; see
@hull2022options and @mcneil2015quantitative.

Let $u_t$ be the daily return and let $\mathcal{F}_{t-1}$ denote the
information available at the end of day $t-1$. The object of interest is

$$
v_t
=
\operatorname{Var}(u_t \mid \mathcal{F}_{t-1}),
$$

where $v_t=\sigma_t^2$ is the conditional variance. The corresponding
conditional volatility is $\sigma_t=\sqrt{v_t}$.

We build the model manually so the bridge remains visible:

$$
\text{returns}
\longrightarrow
\text{GARCH recursion}
\longrightarrow
\text{likelihood}
\longrightarrow
\text{estimated volatility}
\longrightarrow
\text{volatility forecast}.
$$

## Returns and data preparation

The empirical example uses the S&P 500 index from 2015-07-10 to 2020-07-10.
Let $S_t$ be the closing index level. The simple daily return is

$$
u_t
=
\frac{S_t-S_{t-1}}{S_{t-1}}
=
\frac{S_t}{S_{t-1}}-1.
$$

Returns are more appropriate than prices for daily volatility analysis because
they measure relative changes. The square $u_t^2$ measures the size of the
shock regardless of sign.

```{r get-garch-data}
sp500 <- tq_get("^GSPC", from = "2015-07-10", to = "2020-07-10") |>
  select(date, close) |>
  rename(price = close) |>
  mutate(
    return = price / lag(price) - 1
  )

sp500 |>
  select(date, price, return) |>
  slice(c(1:6, (n() - 1):n())) |>
  kable(
    caption = "S&P 500 prices and simple daily returns.",
    digits = 8,
    format.args = list(scientific = FALSE),
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

The first return is missing because the first row has no previous closing
price. We estimate the model after removing that first missing value.

```{r prepare-garch-returns}
returns <- sp500 |>
  filter(!is.na(return)) |>
  select(date, return)

u <- returns$return
long_run_variance <- stats::var(u, na.rm = TRUE)
```

The following plot shows why a conditional volatility model is useful. Large
absolute returns appear in clusters, especially around the 2020 market stress
episode.

```{r}
#| label: fig-garch-returns
#| fig-cap: "S&P 500 daily returns."
returns |>
  ggplot(aes(date, return)) +
  geom_hline(yintercept = 0, color = "gray60", linewidth = 0.4) +
  geom_line(color = "black", linewidth = 0.5) +
  labs(x = "Date", y = "Daily return") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 12)
```

## The GARCH(1,1) model

The GARCH(1,1) model of @bollerslev1986garch extends the ARCH idea of
@engle1982arch by allowing conditional variance to depend on a recent shock and
on its own previous value. Broader treatments of financial time series and
GARCH modeling are given in @tsay2010analysis, @ruppert2011statistics, and
@francq2019garch.

$$
v_t
=
\omega
+
\alpha u_{t-1}^2
+
\beta v_{t-1}.
$$

The model has three parameters:

| Parameter | Interpretation |
|---|---|
| $\omega$ | Baseline variance component. |
| $\alpha$ | Reaction to the most recent squared return. |
| $\beta$ | Persistence from the previous conditional variance. |

The usual restrictions are

$$
\omega>0,
\qquad
\alpha \ge 0,
\qquad
\beta \ge 0,
\qquad
\alpha+\beta<1.
$$

The last restriction gives a finite long-run variance. The sum
$\alpha+\beta$ is especially important: it measures volatility persistence. If
the sum is close to one, shocks fade slowly.

The long-run variance implied by a stationary GARCH(1,1) is

$$
\bar{v}
=
\frac{\omega}{1-\alpha-\beta}.
$$

This identity will be used later for variance targeting and for multi-step
volatility forecasts.

## Conditional variance recursion

The model is recursive. To compute $v_t$, the code needs $u_{t-1}^2$ and
$v_{t-1}$. The first conditional variance must therefore be initialized. To
preserve the source example, we use the shifted initialization from the original
Hull-style material: the first available squared return initializes the second
variance value. The first return is therefore used to start the recursion, and
the likelihood is evaluated from the observations with valid conditional
variance.

The mapping from notation to R is direct:

| Mathematical object | R object |
|---|---|
| $u_t$ | `u[t]` |
| $v_t$ | `v[t]` |
| $\omega$ | `omega` |
| $\alpha$ | `alpha` |
| $\beta$ | `beta` |

```{r garch-recursion}
garch_variance <- function(u, omega, alpha, beta) {
  n <- length(u)
  v <- rep(NA_real_, n)

  if (n >= 2) {
    v[2] <- u[1]^2
  }

  if (n >= 3) {
    for (t in 3:n) {
      v[t] <- omega + alpha * u[t - 1]^2 + beta * v[t - 1]
    }
  }

  v
}
```

The body of the loop mirrors the equation. The code uses `u[t - 1]^2` for the
shock term and `v[t - 1]` for the lagged conditional variance. The shifted
initialization makes the alignment explicit: `u[1]^2` initializes `v[2]`, and
the recursive GARCH update begins at `t = 3`.

## Maximum likelihood estimation

After defining the recursion, we need a statistical criterion for estimating
the parameters. For this teaching example, returns are assumed to be
conditionally normal with zero conditional mean:

$$
u_t \mid \mathcal{F}_{t-1}
\sim
\mathcal{N}(0,v_t).
$$

This assumption gives the log-likelihood contribution

$$
\ell_t^\ast
=
-\log(v_t)
-
\frac{u_t^2}{v_t}.
$$

This is the proportional log-likelihood used in the source example. It omits
constants and multiplies the normal log-likelihood by a positive factor, so the
optimizer searches the same parameter region while preserving the original
reported scale. Since `optim()` minimizes, the R function below returns the
negative of this criterion. Parameter restrictions are handled inside the
function. Invalid parameter vectors receive a very large objective value. For a
more formal treatment of GARCH likelihood, stationarity, and inference, see
@francq2019garch.

The zero-mean assumption keeps the chapter focused on volatility. In a broader
empirical application, $u_t$ could be replaced by residuals from a mean model,
such as a constant-mean or ARMA specification.

Because the shifted recursion leaves the first variance value undefined, the
likelihood below is conditional on the valid variance values produced by the
recursion. The same convention is used for both estimators, so the comparison
remains internally consistent.

```{r garch-likelihood}
nll_garch <- function(par, u) {
  omega <- par[1]
  alpha <- par[2]
  beta <- par[3]

  if (omega <= 0 || alpha < 0 || beta < 0 || (alpha + beta) >= 1) {
    return(1e12)
  }

  v <- garch_variance(u, omega, alpha, beta)
  valid <- !is.na(v) & v > 0

  if (!any(valid)) {
    return(1e12)
  }

  -sum(-log(v[valid]) - (u[valid]^2) / v[valid])
}
```

## Full MLE results

Full maximum likelihood estimates the three GARCH parameters
$(\omega,\alpha,\beta)$. We keep the original numerical strategy: `L-BFGS-B`
handles simple lower and upper bounds, and the objective function penalizes
parameter vectors that violate $\alpha+\beta<1$.

```{r fit-full-garch}
start <- c(omega = 4e-6, alpha = 0.2, beta = 0.7)

fit_full <- optim(
  par = start,
  fn = nll_garch,
  u = u,
  method = "L-BFGS-B",
  lower = c(1e-12, 0, 0),
  upper = c(Inf, 1 - 1e-6, 1 - 1e-6)
)

theta_full <- fit_full$par
omega_full <- theta_full["omega"]
alpha_full <- theta_full["alpha"]
beta_full <- theta_full["beta"]

v_full <- garch_variance(u, omega_full, alpha_full, beta_full)
valid_full <- !is.na(v_full) & v_full > 0

loglik_full <- -nll_garch(theta_full, u)
n_full <- sum(valid_full)
k_full <- 3
aic_full <- -2 * loglik_full + 2 * k_full
bic_full <- -2 * loglik_full + log(n_full) * k_full

ab_full <- alpha_full + beta_full
half_life_full <- ifelse(ab_full < 1, log(0.5) / log(ab_full), NA_real_)

full_summary <- tibble(
  metric = c(
    "omega",
    "alpha",
    "beta",
    "alpha + beta",
    "half-life in trading days",
    "logLik",
    "n_obs",
    "AIC",
    "BIC"
  ),
  value = c(
    omega_full,
    alpha_full,
    beta_full,
    ab_full,
    half_life_full,
    loglik_full,
    n_full,
    aic_full,
    bic_full
  )
)

full_summary |>
  kable(
    caption = "Full maximum likelihood GARCH(1,1) estimates.",
    digits = 6,
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

The parameter $\alpha$ measures the immediate response to a new shock. The
parameter $\beta$ measures persistence. Their sum measures how slowly
conditional variance returns toward its long-run level. The half-life translates
that persistence into trading days:

$$
\text{half-life}
=
\frac{\log(0.5)}{\log(\alpha+\beta)}.
$$

## Variance targeting

Variance targeting changes how $\omega$ is identified. The method fixes a
long-run variance level $V_L$ and uses the stationary GARCH identity

$$
\omega
=
V_L(1-\alpha-\beta).
$$

In this chapter, $V_L$ is the sample variance of returns. The estimation problem
then has two free parameters, $(\alpha,\beta)$, and $\omega$ is calculated from
the long-run variance condition.

```{r fit-vt-garch}
nll_garch_vt <- function(par, u, long_run_variance) {
  alpha <- par[1]
  beta <- par[2]

  if (alpha < 0 || beta < 0 || (alpha + beta) >= 1) {
    return(1e12)
  }

  omega <- long_run_variance * (1 - alpha - beta)
  v <- garch_variance(u, omega, alpha, beta)
  valid <- !is.na(v) & v > 0

  if (!any(valid)) {
    return(1e12)
  }

  -sum(-log(v[valid]) - (u[valid]^2) / v[valid])
}

fit_vt <- optim(
  par = c(alpha = 0.2, beta = 0.7),
  fn = nll_garch_vt,
  u = u,
  long_run_variance = long_run_variance,
  method = "L-BFGS-B",
  lower = c(0, 0),
  upper = c(1 - 1e-6, 1 - 1e-6)
)

alpha_vt <- fit_vt$par["alpha"]
beta_vt <- fit_vt$par["beta"]
omega_vt <- long_run_variance * (1 - alpha_vt - beta_vt)

v_vt <- garch_variance(u, omega_vt, alpha_vt, beta_vt)
valid_vt <- !is.na(v_vt) & v_vt > 0

loglik_vt <- -nll_garch_vt(fit_vt$par, u, long_run_variance)
n_vt <- sum(valid_vt)
k_vt <- 2
aic_vt <- -2 * loglik_vt + 2 * k_vt
bic_vt <- -2 * loglik_vt + log(n_vt) * k_vt

ab_vt <- alpha_vt + beta_vt
half_life_vt <- ifelse(ab_vt < 1, log(0.5) / log(ab_vt), NA_real_)

vt_summary <- tibble(
  metric = c(
    "omega",
    "alpha",
    "beta",
    "alpha + beta",
    "half-life in trading days",
    "logLik",
    "n_obs",
    "AIC",
    "BIC"
  ),
  value = c(
    omega_vt,
    alpha_vt,
    beta_vt,
    ab_vt,
    half_life_vt,
    loglik_vt,
    n_vt,
    aic_vt,
    bic_vt
  )
)

vt_summary |>
  kable(
    caption = "Variance targeting GARCH(1,1) estimates.",
    digits = 6,
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

Variance targeting is a modeling restriction. It reduces the number of
optimized parameters and anchors the long-run variance level. That can improve
numerical stability and communication, provided that the chosen long-run
variance is appropriate for the sample.

## Objective surface and estimator comparison

Variance targeting leaves only two optimized parameters, so the objective
function can be visualized on the $(\alpha,\beta)$ plane. This plot connects
the code to the estimation logic: `optim()` searches over pairs
$(\alpha,\beta)$, evaluates the criterion, and returns the pair with the lowest
objective value. The contour labels use the equivalent proportional
log-likelihood scale, where larger values correspond to lower objective values.
The axis limits keep the same window as the source example, so the solution
remains visible in the relevant region.

```{r}
#| label: fig-garch-vt-contour
#| fig-cap: "Variance targeting objective-function contours."
alpha_grid <- seq(0.22, 0.30, length.out = 100)
beta_grid <- seq(0.66, 0.76, length.out = 100)

objective_surface <- outer(alpha_grid, beta_grid, Vectorize(function(alpha, beta) {
  if (alpha < 0 || beta < 0 || (alpha + beta) >= 0.999) {
    return(NA_real_)
  }

  nll_garch_vt(c(alpha, beta), u = u, long_run_variance = long_run_variance)
}))

loglik_surface <- -objective_surface
contour_levels <- c(10700, 10750, 10800, 10820, 10830, 10835)

contour_lines <- grDevices::contourLines(
  x = beta_grid,
  y = alpha_grid,
  z = t(loglik_surface),
  levels = contour_levels
)

contour_tbl <- bind_rows(lapply(seq_along(contour_lines), function(id) {
  tibble(
    contour_id = id,
    beta = contour_lines[[id]]$x,
    alpha = contour_lines[[id]]$y,
    log_likelihood = contour_lines[[id]]$level
  )
}))

label_candidates <- contour_tbl |>
  filter(
    beta >= 0.662,
    beta <= 0.756,
    alpha >= 0.225,
    alpha <= 0.285
  )

level_tbl <- tibble(
  log_likelihood = contour_levels,
  label_beta_target = c(0.715, 0.730, 0.743, 0.748, 0.752, 0.755)
)

contour_label_tbl <- label_candidates |>
  inner_join(level_tbl, by = "log_likelihood") |>
  group_by(log_likelihood) |>
  slice_min(abs(beta - label_beta_target), n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(label = scales::number(log_likelihood, accuracy = 1, big.mark = ""))

solution_tbl <- tibble(
  beta = beta_vt,
  alpha = alpha_vt
)

contour_tbl |>
  ggplot(aes(beta, alpha)) +
  geom_path(aes(group = contour_id), color = "#1B4E8A", linewidth = 0.5) +
  geom_label(
    data = contour_label_tbl,
    aes(label = label),
    color = "#2F2F2F",
    fill = "white",
    label.size = 0,
    size = 3,
    alpha = 0.9
  ) +
  geom_hline(yintercept = alpha_vt, color = "#B22222", linetype = 2) +
  geom_vline(xintercept = beta_vt, color = "#B22222", linetype = 2) +
  geom_point(
    data = solution_tbl,
    color = "#B22222",
    size = 2.8
  ) +
  coord_cartesian(xlim = c(0.66, 0.76), ylim = c(0.22, 0.30), expand = FALSE) +
  labs(
    x = expression(beta),
    y = expression(alpha),
    caption = "Contour labels show proportional log-likelihood values; higher values imply a lower objective."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

The red point marks the solution returned by `optim()`. The labels on the
contours make the surface easier to read: moving toward higher log-likelihood
values moves toward the criterion that `optim()` minimizes. Long and flat
contours would suggest that several combinations of $\alpha$ and $\beta$
produce similar criterion values.

The two estimators use the same GARCH recursion. The difference is how much
structure they impose on $\omega$.

```{r compare-garch}
comparison_tbl <- bind_rows(
  full_summary |> mutate(model = "Full MLE"),
  vt_summary |> mutate(model = "Variance targeting")
) |>
  pivot_wider(names_from = model, values_from = value) |>
  select(metric, `Full MLE`, `Variance targeting`)

comparison_tbl |>
  kable(
    caption = "Full MLE and variance targeting comparison.",
    digits = 6,
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

The log-likelihood column uses the same proportional criterion as the source
example. AIC and BIC are therefore read on the scale of that convention. They add a
penalty for the number of estimated parameters: Full MLE uses three parameters,
while variance targeting optimizes two parameters and treats $V_L$ as a fixed
plug-in anchor. With that convention, the comparison is about fit and parsimony
together. This style of comparison is common in applied financial time-series
work, where likelihood fit, parameter plausibility, and residual diagnostics
are read together [@tsay2010analysis; @ruppert2011statistics].

The main economic reading comes from $\alpha$, $\beta$, and
$\alpha+\beta$. Similar values across the two estimators indicate that the
long-run variance restriction does not radically change the shock reaction and
persistence estimates in this sample. The larger difference is in
$\alpha+\beta$: Full MLE is closer to the persistence boundary, while variance
targeting gives a shorter half-life. That comparison helps students see that
both estimators share the same recursion, yet the restriction on $\omega$
changes how persistence and long-run variance are balanced.

## Conditional volatility through time and forecasts

The most useful output is the conditional volatility series. We use the
variance targeting estimates for the plot because they are parsimonious and
produce the long-run level explicitly. The same reconstruction can be done with
the Full MLE estimates. Conditional volatility links this chapter to risk
measurement, scenario analysis, and derivatives applications
[@hull2022options; @mcneil2015quantitative].

```{r build-volatility-series}
volatility_tbl <- returns |>
  mutate(
    variance_vt = v_vt,
    volatility_vt = sqrt(variance_vt),
    sample_volatility = sqrt(long_run_variance)
  )

volatility_tbl |>
  select(date, return, variance_vt, volatility_vt) |>
  slice(c(1:6, (n() - 1):n())) |>
  kable(
    caption = "Conditional variance and volatility from variance targeting.",
    digits = 8,
    format.args = list(scientific = FALSE),
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

```{r}
#| label: fig-garch-volatility
#| fig-cap: "S&P 500 conditional daily volatility from GARCH(1,1)."
volatility_tbl |>
  ggplot(aes(date, 100 * volatility_vt)) +
  geom_line(color = "#1B4E8A", linewidth = 0.6) +
  geom_hline(
    aes(yintercept = 100 * sample_volatility),
    color = "#B22222",
    linetype = 2
  ) +
  labs(
    x = "Date",
    y = "Daily volatility",
    caption = "Dashed line: sample daily volatility."
  ) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1, scale = 1)) +
  theme_minimal(base_size = 12)
```

The model assigns more volatility to turbulent periods and less volatility to
calm periods. That is the practical contribution of GARCH relative to a single
sample standard deviation.

After reconstructing the historical conditional volatility path, the final step
is an explicit volatility forecast. Let $T$ be the last observed day. The
one-step-ahead forecast is

$$
\hat{v}_{T+1\mid T}
=
\omega
+
\alpha u_T^2
+
\beta v_T.
$$

For horizons beyond one day, the expected future squared shock is the future
conditional variance. The GARCH(1,1) forecast therefore reverts toward the
long-run variance:

$$
\hat{v}_{T+h\mid T}
=
\bar{v}
+
(\alpha+\beta)^{h-1}
\left(
\hat{v}_{T+1\mid T}-\bar{v}
\right),
\qquad h \ge 1.
$$

```{r garch-volatility-forecast}
last_return <- tail(u, 1)
last_variance <- tail(v_vt, 1)

v_next <- omega_vt + alpha_vt * last_return^2 + beta_vt * last_variance
long_run_vt <- omega_vt / (1 - alpha_vt - beta_vt)

horizon <- 20
forecast_variance <- long_run_vt +
  (alpha_vt + beta_vt)^(seq_len(horizon) - 1) * (v_next - long_run_vt)

forecast_tbl <- tibble(
  horizon = seq_len(horizon),
  variance_forecast = forecast_variance,
  volatility_forecast = sqrt(forecast_variance)
)

forecast_tbl |>
  kable(
    caption = "Twenty-day conditional volatility forecast.",
    digits = 8,
    format.args = list(scientific = FALSE),
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

```{r}
#| label: fig-garch-volatility-forecast
#| fig-cap: "Twenty-day GARCH conditional volatility forecast."
forecast_tbl |>
  ggplot(aes(horizon, 100 * volatility_forecast)) +
  geom_line(color = "#1B4E8A", linewidth = 0.8) +
  geom_point(color = "#1B4E8A", size = 2) +
  geom_hline(
    yintercept = 100 * sqrt(long_run_vt),
    color = "#B22222",
    linetype = 2
  ) +
  labs(
    x = "Forecast horizon in trading days",
    y = "Daily volatility forecast",
    caption = "Dashed line: variance targeting long-run volatility."
  ) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1, scale = 1)) +
  theme_minimal(base_size = 12)
```

This is a risk forecast. Its target is the expected size of future return
variation under the estimated GARCH dynamics; directional return forecasting is
outside this chapter's target.

## Residual diagnostics

After estimating a volatility model, we check whether the standardized
residuals look closer to white noise. The standardized residual is

$$
z_t
=
\frac{u_t}{\sqrt{\hat{v}_t}}.
$$

If GARCH captures the main volatility clustering, the squared standardized
residuals should have less autocorrelation than the squared raw returns.

```{r garch-diagnostics}
diag_tbl <- volatility_tbl |>
  filter(!is.na(variance_vt), variance_vt > 0) |>
  mutate(
    z = return / sqrt(variance_vt),
    z_squared = z^2
  )

lb_z <- stats::Box.test(diag_tbl$z, lag = 20, type = "Ljung-Box")
lb_z_squared <- stats::Box.test(diag_tbl$z_squared, lag = 20, type = "Ljung-Box")

diag_tests <- tibble(
  test = c("Ljung-Box on z_t", "Ljung-Box on z_t squared"),
  statistic = c(as.numeric(lb_z$statistic), as.numeric(lb_z_squared$statistic)),
  p_value = c(lb_z$p.value, lb_z_squared$p.value)
)

diag_tests |>
  kable(
    caption = "Residual diagnostic tests.",
    digits = 4,
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

```{r}
#| label: fig-garch-residual-acf
#| fig-cap: "ACF of squared standardized residuals."
acf_squared <- stats::acf(
  diag_tbl$z_squared,
  plot = FALSE,
  lag.max = 20,
  na.action = na.pass
)

acf_tbl <- tibble(
  lag = as.numeric(acf_squared$lag[-1]),
  acf = as.numeric(acf_squared$acf[-1])
)

acf_tbl |>
  ggplot(aes(lag, acf)) +
  geom_col(width = 0.08, fill = "#1B4E8A") +
  geom_hline(
    yintercept = c(-1.96 / sqrt(nrow(diag_tbl)), 1.96 / sqrt(nrow(diag_tbl))),
    linetype = 2,
    color = "#B22222"
  ) +
  labs(x = "Lag", y = "ACF") +
  theme_minimal(base_size = 12)
```

The diagnostic is a model audit. A small p-value for the squared standardized
residuals would suggest remaining volatility structure. An ACF with smaller
remaining spikes supports the interpretation that the GARCH recursion absorbed
much of the conditional heteroskedasticity.

## Common implementation mistakes

A manual GARCH implementation is useful because it makes mistakes visible. The
most important checks are:

| Check | Why it is important |
|---|---|
| Lag alignment | The recursion must use $u_{t-1}^2$ and $v_{t-1}$ to compute $v_t$. |
| Scale consistency | Returns should stay in decimal units during estimation. Percent units can be used for plots. |
| Parameter restrictions | $\omega>0$, $\alpha\ge0$, $\beta\ge0$, and $\alpha+\beta<1$ keep the model interpretable. |
| Starting values | Different feasible starting values should lead to similar estimates in a stable problem. |
| Diagnostics | AIC and BIC should be read together with parameter plausibility and residual checks. |

These checks are part of the model workflow. They protect the correspondence
between the equation, the code, and the reported result.

## Conclusion and extensions

This chapter built a GARCH(1,1) volatility forecasting workflow from first
principles. The data are daily S&P 500 returns. The model updates conditional
variance with the most recent squared return and the previous conditional
variance. The likelihood turns that recursion into an estimable criterion, and
the estimated parameters produce a conditional volatility series.

Full MLE and variance targeting use the same recursion. Full MLE estimates
$(\omega,\alpha,\beta)$ freely. Variance targeting estimates $(\alpha,\beta)$
and constructs $\omega$ from a long-run variance condition. The comparison
therefore teaches a broader modeling lesson: estimation is also a choice about
which restrictions are reasonable for the question.

The volatility forecast is the central forecasting object in this chapter. It
summarizes expected future risk; directional return forecasting is outside the
model target. In an applied forecasting book, that distinction is important:
forecasting can target a level, a mean, a probability, or a variance. Here the
target is dynamic risk.

The natural extensions follow from the assumptions made here. GARCH(1,1) is a
starting point. Common extensions include asymmetric volatility models such as
EGARCH and GJR-GARCH, heavy-tailed conditional distributions such as Student-t
innovations, and multivariate models such as DCC-GARCH. These extensions are
treated in more advanced references such as @francq2019garch, @tsay2010analysis,
and @mcneil2015quantitative. The core logic remains the same: define a
conditional variance process, estimate it with an explicit criterion, diagnose
the residuals, and evaluate whether the forecast target is useful for the
applied problem.
