# GARCH volatility forecasting in R

```{r}
# This removes all items in environment.
rm(list = ls())

library(tidyquant)
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(knitr)
library(kableExtra)
library(scales)
```

## Volatility as a forecasting problem

An investor observing a quiet market today may expect relatively small price
movements tomorrow. After several turbulent sessions, the same investor may
revise that assessment even without changing the expected direction of the
market. This chapter develops a model for that changing assessment of risk.
Where the previous chapters forecast the level of a monthly sales series,
we now ask how much a daily financial return is likely to vary, given what
has already happened in the market. The implementation builds on the GARCH
example in @lozano2025garchvideo and the risk-management applications discussed
by @hull2022options and @mcneil2015quantitative.

The starting observation is that large financial returns tend to occur close
together in time. A turbulent session is often followed by further large
movements, while quiet periods can persist for many days. A standard deviation
calculated from the entire sample averages over both kinds of period. To
describe the risk faced on a particular date, we instead allow variance to
change as new returns are observed. A GARCH model provides a rule for making
these updates and carrying them forward into a forecast.

Let $u_t$ be the daily return and let $\mathcal{F}_{t-1}$ denote the
information available at the end of day $t-1$. The object of interest is

$$
v_t
=
\operatorname{Var}(u_t \mid \mathcal{F}_{t-1}).
$$

The conditioning information in this expression fixes the timing of the
question: $v_t$ is the variance assigned to day $t$ using information available
at the end of day $t-1$. Its square root, $\sigma_t=\sqrt{v_t}$, is the
conditional volatility, expressed in the same units as the return. By
constructing the model step by step, we can follow how the data lead to these
daily risk estimates and eventually to a forecast:

$$
\text{returns}
\longrightarrow
\text{GARCH recursion}
\longrightarrow
\text{likelihood}
\longrightarrow
\text{estimated volatility}
\longrightarrow
\text{volatility forecast}.
$$

The first arrow begins with a measurement choice. We convert index levels to
returns so that a price movement is measured relative to the amount invested.
Those returns become inputs to the GARCH recursion, which calculates today's
variance from yesterday's squared return and variance. At first, however, the
parameters of that rule are unknown. Different choices produce different
variance paths, and we need a way to decide which path best accounts for the
observations. The likelihood supplies that comparison by measuring how
plausible the observed returns are under each proposed path. Estimation
therefore moves repeatedly between the recursion and the likelihood until
the numerical search finds the best-fitting parameters.

Once those parameters have been selected, the recursion gives a fitted
variance for each date; taking square roots produces the estimated volatility
series in the diagram. Extending this series into the future requires one
further change. Historical updates can use returns that have already occurred,
whereas future updates depend on shocks that have yet to arrive. Replacing
their squared values by the expectations implied by the model gives the
forecast. The distinction between observed shocks and expected future shocks
will explain both the shape of that forecast and how it changes when new data
become available. Before developing the updating rule, we need to construct
the returns on which it operates.

## Returns and data preparation

Our application uses the S&P 500 index, with prices requested from 2015-07-10
to a download cutoff of 2020-07-10. Because the index level changes over time,
a movement of a given number of points can represent very different changes
in the value of an investment. We account for this by dividing the change
in the closing level, $S_t-S_{t-1}$, by the preceding close, $S_{t-1}$.
The resulting simple daily return is

$$
u_t
=
\frac{S_t-S_{t-1}}{S_{t-1}}
=
\frac{S_t}{S_{t-1}}-1.
$$

For example, a rise of 10 points from 100 to 110 gives a 10% return, while
the same rise from 1,000 to 1,010 gives a 1% return. The absolute changes are
identical, but the first movement is ten times as large relative to the
starting value. Measuring returns therefore lets us compare risk across
assets with different prices and across periods in which the same index
trades at different levels. Throughout the calculations, we express these
returns as decimals, so that 0.01 consistently represents 1%.

Working with returns also brings the statistical measurement closer to the
financial question. Price levels often trend and retain the accumulated
effects of past shocks; their dispersion over a long sample can consequently
reflect years of growth as well as short-run uncertainty. Daily returns
isolate the relative movements whose variability we want to forecast. This
transformation still leaves an important feature to explain: the size of
those movements can change markedly from one period to another. GARCH models
that changing conditional variance. Whether the returns are stationary and
whether this particular model captures their behavior remain questions to
assess from the data.

```{r get-garch-data}
sp500 <- tq_get("^GSPC", from = "2015-07-10", to = "2020-07-10") |>
  select(date, close) |>
  rename(price = close) |>
  mutate(
    return = price / lag(price) - 1
  )

sp500 |>
  select(date, price, return) |>
  slice(c(1:6, (n() - 1):n())) |>
  kable(
    caption = "S&P 500 prices and simple daily returns.",
    col.names = c("Date", "S&P 500 level", "\\(u_t\\)"),
    escape = FALSE,
    digits = 8,
    format.args = list(scientific = FALSE),
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

The table contains prices from `r min(sp500$date)` through
`r max(sp500$date)`. Its first row has no return because calculating a change
requires a preceding close. We remove that initial missing value when
constructing the return vector, while retaining the original prices for the
calculation tables below. The resulting return sample will serve throughout
as our training data. Later, we will evaluate forecasts using subsequent
observations that have played no part in estimating the parameters.

```{r prepare-garch-returns}
returns <- sp500 |>
  filter(!is.na(return)) |>
  select(date, return)

u <- returns$return
long_run_variance <- stats::var(u, na.rm = TRUE)
```

Plotting these returns reveals why a single measure of dispersion would be
an incomplete description of risk. Around the 2020 market stress episode,
large positive and negative movements appear close together, in contrast to
the much smaller movements observed during quieter periods. An updating rule
for variance should be able to respond to this clustering.

```{r}
#| label: fig-garch-returns
#| fig-cap: "S&P 500 daily returns."
returns |>
  ggplot(aes(date, return)) +
  geom_hline(yintercept = 0, color = "gray60", linewidth = 0.4) +
  geom_line(color = "black", linewidth = 0.5) +
  labs(x = "Date", y = expression("Daily Return " * u[t])) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal(base_size = 12)
```

## The GARCH(1,1) model

To describe the clustering visible in the return plot, we let a large movement
raise the variance assigned to the following day and allow some of that
increase to carry forward. The GARCH(1,1) model of @bollerslev1986garch
implements both ideas in one equation. It extends the ARCH model of
@engle1982arch by including the preceding conditional variance alongside the
preceding squared shock:

$$
v_t
=
\omega
+
\alpha u_{t-1}^2
+
\beta v_{t-1}.
$$

The intercept and the weights on these two sources of information are the
three parameters we will estimate:

| Parameter | Interpretation |
|---|---|
| $\omega$ | Baseline variance component. |
| $\alpha$ | Reaction to the most recent squared return. |
| $\beta$ | Persistence from the previous conditional variance. |

The use of a squared return follows from the question we are asking. A large
gain and a large loss both indicate substantial variation, even though their
directions differ. Under the zero-conditional-mean assumption used here,
$u_t$ is also the unexpected return, or shock, and $u_t^2$ measures its
magnitude without retaining its sign. Returns of 2% and -2%, for instance,
both give $u_t^2=0.0004$, four times the squared value of a 1% return.
Through $\alpha u_{t-1}^2$, a large movement in either direction can therefore
raise the next variance estimate. A single squared return is a noisy indication
of risk, however, so the equation also draws on the previous variance instead
of relying entirely on that one observation. The equal response to positive
and negative shocks of the same size is a feature of this symmetric model.

The preceding variance enters through $\beta v_{t-1}$, allowing the model to
retain some of its earlier assessment of risk. This provides one source of
persistence, but the squared-return term adds another: when variance is high,
larger squared returns are also expected, and these feed back into subsequent
updates. Taking expectations over future shocks combines the two effects
into $\alpha+\beta$. If this sum is close to one, an increase in variance
is expected to fade slowly. Along an observed path, further shocks can
interrupt that decline, so persistence describes expected propagation rather
than a smooth recovery that must occur after every turbulent day.

For this updating rule to produce meaningful variances and a finite long-run
level, we impose the restrictions

$$
\omega>0,
\qquad
\alpha \ge 0,
\qquad
\beta \ge 0,
\qquad
\alpha+\beta<1.
$$

The first three conditions protect the sign of the calculated variance. With
a positive intercept and nonnegative weights, the terms in the recursion
cannot combine to produce a negative value. Allowing $\omega<0$ would remove
that protection on a quiet day, when the other contributions are small;
allowing $\alpha<0$ would make a sufficiently large squared shock push the
next variance below zero. A negative $\beta$ would create a similar problem
when the preceding variance is large. We permit either weight to equal zero,
which simply removes its channel, but require the intercept to remain strictly
positive. With $\alpha+\beta<1$, setting the intercept to zero would imply
a degenerate zero long-run variance.

Positivity alone does not determine whether the process settles around a
finite average level. To examine that question, take unconditional
expectations in the recursion. Under the zero-mean assumption, the average
squared return equals the average conditional variance, denoted by $\bar v$.
The equation then becomes $\bar v=\omega+(\alpha+\beta)\bar v$, which gives

$$
\bar{v}
=
\frac{\omega}{1-\alpha-\beta}.
$$

This expression explains the restriction on the sum of the weights. For
$\omega>0$, a positive, finite long-run variance requires $\alpha+\beta<1$.
At a sum of one, expected variance fails to return to a finite level, and
above one its expected dynamics are explosive. The finite-second-moment
condition is the one needed for our variance-targeting estimator and
mean-reverting forecasts below. Strict stationarity is a distinct, more
general issue; treatments of that distinction and of financial time-series
modeling more broadly can be found in @francq2019garch, @tsay2010analysis,
and @ruppert2011statistics.

## Conditional variance recursion

The equation gives us a rule for one day, but applying it to the sample
requires a sequence of calculations. Once we have obtained $v_t$ from
$u_{t-1}^2$ and $v_{t-1}$, that new variance becomes an input to the
calculation of $v_{t+1}$. We combine it with $u_t^2$ and continue forward
in the same way. This dependence of each calculation on the preceding result
is what makes the model recursive.

To begin the sequence, we need a variance value that the data do not directly
provide. We use the first available squared return to initialize the variance
for the following return date. Counting from the first row of prices makes
the timing clear: day 1 supplies the initial close, day 2 supplies the first
return, and day 3 receives the initialized variance $v_3=u_2^2$. The full
GARCH update can then begin on day 4, using the return and initialized variance
from day 3. Because a return and a positive variance are available together
from day 3 onward, those are also the observations that enter the likelihood.

The R function implements this sequence using the following objects:

| Mathematical object | R object | Description |
|---|---|---|
| $u_t$ | `u[t]` | Daily return, also the shock under the zero-mean assumption. |
| $v_t$ | `v[t]` | Conditional variance assigned to that return date. |
| $\omega$ | `omega` | Positive intercept in the variance recursion. |
| $\alpha$ | `alpha` | Weight on the preceding squared return. |
| $\beta$ | `beta` | Weight on the preceding conditional variance. |

In this mapping, `t` indexes the return vector after the first missing return
has been removed. Its first element corresponds to price-row day 2. The
calculation tables below retain the original price-row day numbers so that
the initial missing values remain visible.

```{r garch-recursion}
garch_variance <- function(u, omega, alpha, beta) {
  n <- length(u)
  v <- rep(NA_real_, n)

  if (n >= 2) {
    v[2] <- u[1]^2
  }

  if (n >= 3) {
    for (t in 3:n) {
      v[t] <- omega + alpha * u[t - 1]^2 + beta * v[t - 1]
    }
  }

  v
}
```

The function first fills a variance vector with missing values and sets
`v[2]` equal to `u[1]^2`. The loop can then begin at `t = 3`, using
`u[t - 1]^2` and `v[t - 1]` in exactly the order required by the equation.
These indices refer to the shortened return vector, so this first loop
iteration corresponds to day 4 in the original price data. We now have a
function that produces a complete variance path for any admissible parameter
vector. The remaining question is how to choose that vector from the data.

## Maximum likelihood estimation

Suppose we try a parameter vector that assigns very low variance to a period
of large returns. Those observations would be difficult to explain under the
proposed model. Raising the variance would make them more plausible, but a
rule that assigns high variance on every date would also fit quiet periods
poorly. Maximum likelihood formalizes this comparison by evaluating how much
probability density the model assigns near each observed return, conditional
on the information preceding it. We use density because returns are continuous:
the probability of any one exact value is zero, while the probability of a
small interval around it depends on the density there.

To make this comparison numerical, we need a distribution whose spread is
determined by the GARCH variance. We assume

$$
u_t \mid \mathcal{F}_{t-1}
\sim
\mathcal{N}(0,v_t),
$$

so that a small $v_t$ concentrates the normal density around zero and a larger
$v_t$ spreads it over a wider range of possible returns. The zero mean lets
us focus on this changing spread; in an application with a separate mean
model, such as a constant or an ARMA equation, its residuals would take the
place of $u_t$ in the variance calculation.

Multiplying the conditional densities over the sample gives the likelihood
of the observed sequence, conditional on our initialization. We work with its
logarithm because it turns this product into a sum without changing which
parameter vector is preferred. This also avoids the numerical difficulty
of multiplying many small densities. For estimation, it is convenient to
begin with the following proportional daily contribution:

$$
\ell_t^\ast
=
-\log(v_t)
-
\frac{u_t^2}{v_t}.
$$

The trade-off described above is visible in the two terms. If a large return
occurs when $v_t$ is small, the ratio $u_t^2/v_t$ becomes large and reduces
the contribution through its negative sign. Increasing $v_t$ eases that
penalty, but also lowers $-\log(v_t)$, so making every variance arbitrarily
large cannot solve the fitting problem. The optimizer must balance these
effects across the sample using one common set of GARCH parameters.

Adding the daily contributions produces $\ell^*=\sum_t\ell_t^\ast$.
The asterisk reminds us that this score has been simplified for estimation.
To report the complete Gaussian log-likelihood, we restore the normal-density
constant and the factor of one half:

$$
\ell
=
-\frac{1}{2}
\sum_t
\left[
\log(2\pi)
+
\log(v_t)
+
\frac{u_t^2}{v_t}
\right].
$$

Both expressions use the same variance path and the same dates. Their
dependence on $(\omega,\alpha,\beta)$ comes through the variances generated
by the recursion; the additional constant contains no unknown parameter.
Consequently, if $n$ is the number of included observations, we can obtain
the full Gaussian total directly from the proportional total:

$$
\ell
=
\frac{\ell^\ast}{2}
-
\frac{n}{2}\log(2\pi).
$$

Dividing every candidate score by two and subtracting the same constant
preserves its ordering. The parameters that maximize $\ell^*$ therefore also
maximize $\ell$, even though the reported totals differ. Either total can be
positive because a density can exceed one; the relevant comparison is which
parameter vector gives the higher value on the same data and scale.

That equivalence is enough for parameter estimation, but information criteria
combine the likelihood with a penalty whose scale is fixed. The formulas

$$
AIC=-2\ell+2k,
\qquad
BIC=-2\ell+k\log(n)
$$

use $k$ to count the estimated quantities entering the penalty. Substituting
$\ell^*$ for $\ell$ would double the relative weight assigned to fit while
leaving that penalty unchanged. We therefore use the complete Gaussian value
for both criteria. Although its constant cancels in comparisons on identical
observations, retaining it also makes the reported likelihood conform to
the usual definition.

The code reflects this distinction. The estimation functions `nll_garch()`
and `nll_garch_vt()` return $-\ell^*$ because R's `optim()` searches for a
minimum. After estimation, a returned value called `nll` can be put on the
Gaussian scale using

```r
loglik_gaussian <- -nll / 2 - n / 2 * log(2 * pi)
```

We also calculate the Gaussian likelihood directly from the returns and
fitted variances in `gaussian_loglik()`. Agreement between that calculation
and the conversion provides a check on the implementation. Both use only
dates with valid variances, since the shifted initialization leaves the
first return without an associated variance value. Keeping this convention
identical for the two estimators ensures that differences in fit come from
their parameters rather than their observation sets.

During the search, a vector that violates the parameter restrictions receives
a very large objective value. This keeps such a vector from being accepted
as a better fit simply because the variance calculation has become invalid.
The resulting conditional-likelihood procedure is developed more formally
in @francq2019garch; here we can trace each of its steps in the two functions
below.

```{r garch-likelihood}
gaussian_loglik <- function(u, v) {
  valid <- is.finite(u) & is.finite(v) & v > 0

  if (!any(valid)) {
    stop("No valid observations are available for the Gaussian likelihood.")
  }

  list(
    value = -0.5 * sum(
      log(2 * pi) + log(v[valid]) + (u[valid]^2) / v[valid]
    ),
    n = sum(valid),
    valid = valid
  )
}

nll_garch <- function(par, u) {
  omega <- par[1]
  alpha <- par[2]
  beta <- par[3]

  if (omega <= 0 || alpha < 0 || beta < 0 || (alpha + beta) >= 1) {
    return(1e12)
  }

  v <- garch_variance(u, omega, alpha, beta)
  valid <- !is.na(v) & v > 0

  if (!any(valid)) {
    return(1e12)
  }

  -sum(-log(v[valid]) - (u[valid]^2) / v[valid])
}
```

## Full MLE results

We first allow the likelihood to choose all three parameters jointly.
Returning to the model equation makes clear what is being estimated:

$$
v_t=\omega+\alpha u_{t-1}^2+\beta v_{t-1}.
$$

For each trial choice of $(\omega,\alpha,\beta)$, the recursion produces a
variance path and the likelihood assigns it a score. We use L-BFGS-B to search
for the vector with the lowest negative score. Its name refers to the
limited-memory Broyden-Fletcher-Goldfarb-Shanno algorithm with bound constraints.
As a quasi-Newton method, it learns about the objective's curvature from
changes in its slope and uses that information to choose the next search
direction. It keeps a limited record of recent steps, avoiding the need to
construct an exact second-derivative matrix at every iteration. In our call
to `optim()`, the slopes themselves are approximated numerically because we
do not supply a derivative function.

Its bound constraints are useful here because the parameters cannot take
arbitrary values. We specify a positive lower bound for $\omega$ and
nonnegative lower bounds for $\alpha$ and $\beta$. However, a separate bound
on each coefficient still permits some combinations with
$\alpha+\beta\ge1$, so the penalty in the objective handles that joint
restriction. The reported convergence code tells us whether the search met
its stopping rule. We will assess the solution further by checking its
admissibility and comparing its likelihood with the variance-targeting fit.

```{r fit-full-garch}
start <- c(omega = 4e-6, alpha = 0.2, beta = 0.7)

fit_full <- optim(
  par = start,
  fn = nll_garch,
  u = u,
  method = "L-BFGS-B",
  lower = c(1e-12, 0, 0),
  upper = c(Inf, 1 - 1e-6, 1 - 1e-6),
  control = list(
    parscale = c(1e-6, 0.1, 0.1),
    factr = 1e7,
    pgtol = 1e-10
  )
)

if (fit_full$convergence != 0) {
  stop("The scaled Full MLE optimization did not converge.")
}

theta_full <- fit_full$par
omega_full <- theta_full["omega"]
alpha_full <- theta_full["alpha"]
beta_full <- theta_full["beta"]

v_full <- garch_variance(u, omega_full, alpha_full, beta_full)
likelihood_full <- gaussian_loglik(u, v_full)
valid_full <- likelihood_full$valid
n_full <- likelihood_full$n
loglik_full <- likelihood_full$value
nll_full <- nll_garch(theta_full, u)
loglik_full_from_conversion <-
  -nll_full / 2 - n_full / 2 * log(2 * pi)

stopifnot(isTRUE(all.equal(
  loglik_full,
  loglik_full_from_conversion,
  tolerance = 1e-10
)))

k_full <- 3
aic_full <- -2 * loglik_full + 2 * k_full
bic_full <- -2 * loglik_full + log(n_full) * k_full

ab_full <- alpha_full + beta_full
half_life_full <- ifelse(ab_full < 1, log(0.5) / log(ab_full), NA_real_)

full_summary <- tibble(
  metric = c(
    "omega",
    "alpha",
    "beta",
    "alpha + beta",
    "half-life in trading days",
    "logLik",
    "n_obs",
    "AIC",
    "BIC"
  ),
  value = c(
    omega_full,
    alpha_full,
    beta_full,
    ab_full,
    half_life_full,
    loglik_full,
    n_full,
    aic_full,
    bic_full
  )
)

full_summary |>
  kable(
    caption = "Full maximum likelihood GARCH(1,1) estimates with Gaussian log-likelihood.",
    digits = 6,
    col.names = c("Metric", "Value"),
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

The estimates in the table also explain the scaling used in the optimization
call. The intercept is of order $10^{-6}$, while the two weights are of order
tenths, so a numerically sensible step in one coordinate can be far too large
or too small in another. The `parscale` argument allows the search to account
for these different magnitudes. It ends with convergence code
`r fit_full$convergence`, and `gaussian_loglik()` then evaluates the fitted
path on the scale used for `logLik`, AIC, and BIC. Inserting the coefficients
back into the model gives the daily updating rule

$$
\hat v_t=
`r sprintf("%.10f", omega_full)`
+ `r sprintf("%.8f", alpha_full)`\,u_{t-1}^2
+ `r sprintf("%.8f", beta_full)`\,\hat v_{t-1}.
$$

We can now follow this rule through the first few observations. In the table
below, day $i$ counts from the first price in `sp500`, so the return on day 2
is `u[1]` and the initialized variance on day 3 is `v_full[2]`. The dashes
show where an earlier observation is still missing. Once both the current
return and its variance are available, they give the daily contribution in
the last column, $\ell_i^*=-\log(\hat v_i)-u_i^2/\hat v_i$. Adding these
contributions over all defined rows, including those omitted from the display,
reproduces the total score optimized above. Returns are shown in decimals,
with variances in the corresponding squared units.

In the numerical substitutions below, equality signs describe the calculation
carried out at full precision. Displayed inputs and results are rounded, so
recalculating from the printed digits may differ in the last decimal places.

```{r garch-daily-calculation}
garch_calculation <- function(prices, variances) {
  daily <- prices |>
    mutate(day = row_number(), variance = c(NA_real_, variances)) |>
    select(date, day, price, return, variance)
  daily$contribution <- NA_real_
  valid <- is.finite(daily$return) & is.finite(daily$variance) & daily$variance > 0
  daily$contribution[valid] <- -log(daily$variance[valid]) -
    daily$return[valid]^2 / daily$variance[valid]
  daily
}

compact_garch_calculation <- function(daily) {
  display <- daily |>
    mutate(
      date = as.character(date), day = as.character(day),
      price = sprintf("%.2f", price),
      return = ifelse(is.na(return), "--", sprintf("%.10f", return)),
      variance = ifelse(is.na(variance), "--", sprintf("%.10f", variance)),
      contribution = ifelse(is.na(contribution), "--", sprintf("%.6f", contribution))
    )
  omitted <- as_tibble(setNames(rep(list("..."), ncol(display)), names(display)))
  bind_rows(head(display, 6), omitted, tail(display, 2))
}

full_daily <- garch_calculation(sp500, v_full)
stopifnot(isTRUE(all.equal(
  sum(full_daily$contribution, na.rm = TRUE), -nll_full, tolerance = 1e-10
)))
compact_garch_calculation(full_daily) |>
  kable(
    caption = "Daily GARCH(1,1) variance and likelihood calculations: full maximum likelihood.",
    col.names = c("Date", "Day", "S&P 500", "\\(u_i\\)", "\\(\\hat v_i\\)", "\\(\\ell_i^*\\)"),
    escape = FALSE, row.names = FALSE, align = c("l", rep("r", 5))
  ) |>
  kable_styling(full_width = FALSE, latex_options = "HOLD_position")
```

Day 3 is initialized as
$\hat v_3=u_2^2=(`r sprintf("%.10f", full_daily$return[2])`)^{2}
= `r sprintf("%.10f", full_daily$variance[3])`$.
The parameters first enter the variance calculation on day 4. Reading the
preceding return and variance from the table gives

$$
\begin{aligned}
\hat v_4
&=\hat\omega+\hat\alpha u_3^2+\hat\beta\hat v_3\\
&=`r sprintf("%.10f", omega_full)`
+`r sprintf("%.8f", alpha_full)`(`r sprintf("%.10f", full_daily$return[3])`)^2\\
&\quad+`r sprintf("%.8f", beta_full)`(`r sprintf("%.10f", full_daily$variance[3])`)\\
&= `r sprintf("%.10f", full_daily$variance[4])`.
\end{aligned}
$$

We then carry this newly calculated variance forward one day:

$$
\begin{aligned}
\hat v_5
&=\hat\omega+\hat\alpha u_4^2+\hat\beta\hat v_4\\
&=`r sprintf("%.10f", omega_full)`
+`r sprintf("%.8f", alpha_full)`(`r sprintf("%.10f", full_daily$return[4])`)^2\\
&\quad+`r sprintf("%.8f", beta_full)`(`r sprintf("%.10f", full_daily$variance[4])`)\\
&= `r sprintf("%.10f", full_daily$variance[5])`.
\end{aligned}
$$

The resulting values reproduce the fifth column for days 4 and 5, subject
to the rounding used in the displayed substitutions; R retains full precision
throughout. Notice that each update uses only the preceding row. The current
return enters afterwards, when we evaluate the likelihood contribution for
the variance already assigned to that date. This is how the same calculation
can serve both estimation and forecasting without using a return before it
has been observed.

The estimated weights tell us how long the effect of a large movement is
expected to remain. While $\alpha$ determines the immediate response and
$\beta$ carries forward the preceding variance, their sum governs the
expected decay toward the long-run level. To express that persistence in
trading days, we calculate the half-life:

$$
\text{half-life}
=
\frac{\log(0.5)}{\log(\alpha+\beta)}.
$$

With an estimated persistence of `r round(ab_full, 4)`, it takes approximately
`r round(half_life_full, 1)` trading days for an initial excess variance to
fall by half in expectation. For instance, a variance initially 0.00010 above
its long-run level would be expected to lie about 0.00005 above it after one
half-life. The quantity being halved is the deviation from the long-run
variance; total volatility is obtained by taking the square root of the full
variance at each date. New shocks can alter the actual path, but the half-life
still gives a useful financial reading of the fitted dynamics: a longer
half-life means that an episode of high risk continues to influence forecasts
for longer. We next examine how these dynamics change when estimation is
anchored to a specified long-run variance.

## Variance targeting

The first estimator allowed all three parameters to adjust to the likelihood.
Variance targeting retains the same GARCH(1,1) equation,

$$
v_t=\omega+\alpha u_{t-1}^2+\beta v_{t-1},
$$

but ties its long-run variance to a value $V_L$ chosen before optimizing the
weights. The stationary-variance identity then determines the intercept:

$$
\omega
=
V_L(1-\alpha-\beta).
$$

We use the training sample variance of returns for $V_L$. For each proposed
pair $(\alpha,\beta)$, the equation above supplies $\omega$, after which the
same recursion and likelihood can be evaluated. The search therefore has two
free coordinates. Its reduction in dimension comes from anchoring the model's
long-run level to the sample estimate, while allowing the likelihood to choose
how quickly the variance responds and how persistent those responses are.

```{r fit-vt-garch}
nll_garch_vt <- function(par, u, long_run_variance) {
  alpha <- par[1]
  beta <- par[2]

  if (alpha < 0 || beta < 0 || (alpha + beta) >= 1) {
    return(1e12)
  }

  omega <- long_run_variance * (1 - alpha - beta)
  v <- garch_variance(u, omega, alpha, beta)
  valid <- !is.na(v) & v > 0

  if (!any(valid)) {
    return(1e12)
  }

  -sum(-log(v[valid]) - (u[valid]^2) / v[valid])
}

fit_vt <- optim(
  par = c(alpha = 0.2, beta = 0.7),
  fn = nll_garch_vt,
  u = u,
  long_run_variance = long_run_variance,
  method = "L-BFGS-B",
  lower = c(0, 0),
  upper = c(1 - 1e-6, 1 - 1e-6)
)

if (fit_vt$convergence != 0) {
  stop("The variance targeting optimization did not converge.")
}

alpha_vt <- fit_vt$par["alpha"]
beta_vt <- fit_vt$par["beta"]
omega_vt <- long_run_variance * (1 - alpha_vt - beta_vt)

v_vt <- garch_variance(u, omega_vt, alpha_vt, beta_vt)
likelihood_vt <- gaussian_loglik(u, v_vt)
valid_vt <- likelihood_vt$valid
n_vt <- likelihood_vt$n
loglik_vt <- likelihood_vt$value
nll_vt <- nll_garch_vt(fit_vt$par, u, long_run_variance)
loglik_vt_from_conversion <-
  -nll_vt / 2 - n_vt / 2 * log(2 * pi)

stopifnot(isTRUE(all.equal(
  loglik_vt,
  loglik_vt_from_conversion,
  tolerance = 1e-10
)))

theta_vt_in_full <- c(
  omega = unname(omega_vt),
  alpha = unname(alpha_vt),
  beta = unname(beta_vt)
)
nll_vt_from_full <- nll_garch(theta_vt_in_full, u)

stopifnot(isTRUE(all.equal(
  nll_vt,
  nll_vt_from_full,
  tolerance = 1e-10
)))

if (!identical(which(valid_full), which(valid_vt))) {
  stop("Full MLE and variance targeting use different likelihood observations.")
}

if (loglik_full + 1e-8 < loglik_vt) {
  stop("Full MLE has a lower Gaussian log-likelihood than variance targeting.")
}

k_vt <- 2
aic_vt <- -2 * loglik_vt + 2 * k_vt
bic_vt <- -2 * loglik_vt + log(n_vt) * k_vt

ab_vt <- alpha_vt + beta_vt
half_life_vt <- ifelse(ab_vt < 1, log(0.5) / log(ab_vt), NA_real_)

vt_summary <- tibble(
  metric = c(
    "omega",
    "alpha",
    "beta",
    "alpha + beta",
    "half-life in trading days",
    "logLik",
    "n_obs",
    "AIC",
    "BIC"
  ),
  value = c(
    omega_vt,
    alpha_vt,
    beta_vt,
    ab_vt,
    half_life_vt,
    loglik_vt,
    n_vt,
    aic_vt,
    bic_vt
  )
)

vt_summary |>
  kable(
    caption = "Variance targeting GARCH(1,1) estimates with Gaussian log-likelihood.",
    digits = 6,
    col.names = c("Metric", "Value"),
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

With $V_L=`r sprintf("%.10f", long_run_variance)`$, the optimized weights
imply $\hat\omega=V_L(1-\hat\alpha-\hat\beta)
=`r sprintf("%.10f", omega_vt)`$. Although the table reports three
coefficients, only $\alpha$ and $\beta$ have been chosen directly by the
optimizer. Their values and the long-run input together give the fitted rule

$$
\hat v_t=
`r sprintf("%.10f", omega_vt)`
+ `r sprintf("%.8f", alpha_vt)`\,u_{t-1}^2
+ `r sprintf("%.8f", beta_vt)`\,\hat v_{t-1}.
$$

Applying this rule to the same price and return sequence lets us see where
the two procedures begin to differ. Their initialized variance on day 3 is
identical, since it depends only on the first available return. Differences
first appear on day 4, when the fitted coefficients enter the recursive
update. Keeping the dates and initialization fixed makes the following table
directly comparable with its full maximum likelihood counterpart.

```{r vt-daily-calculation}
vt_daily <- garch_calculation(sp500, v_vt)
stopifnot(isTRUE(all.equal(
  sum(vt_daily$contribution, na.rm = TRUE), -nll_vt, tolerance = 1e-10
)))
compact_garch_calculation(vt_daily) |>
  kable(
    caption = "Daily GARCH(1,1) variance and likelihood calculations: variance targeting.",
    col.names = c("Date", "Day", "S&P 500", "\\(u_i\\)", "\\(\\hat v_i\\)", "\\(\\ell_i^*\\)"),
    escape = FALSE, row.names = FALSE, align = c("l", rep("r", 5))
  ) |>
  kable_styling(full_width = FALSE, latex_options = "HOLD_position")
```

Using the variance-targeting coefficients, the first recursive update is

$$
\begin{aligned}
\hat v_4
&=\hat\omega+\hat\alpha u_3^2+\hat\beta\hat v_3\\
&=`r sprintf("%.10f", omega_vt)`
+`r sprintf("%.8f", alpha_vt)`(`r sprintf("%.10f", vt_daily$return[3])`)^2\\
&\quad+`r sprintf("%.8f", beta_vt)`(`r sprintf("%.10f", vt_daily$variance[3])`)\\
&= `r sprintf("%.10f", vt_daily$variance[4])`.
\end{aligned}
$$

For day 5, we use the return on day 4 and the variance just obtained:

$$
\begin{aligned}
\hat v_5
&=\hat\omega+\hat\alpha u_4^2+\hat\beta\hat v_4\\
&=`r sprintf("%.10f", omega_vt)`
+`r sprintf("%.8f", alpha_vt)`(`r sprintf("%.10f", vt_daily$return[4])`)^2\\
&\quad+`r sprintf("%.8f", beta_vt)`(`r sprintf("%.10f", vt_daily$variance[4])`)\\
&= `r sprintf("%.10f", vt_daily$variance[5])`.
\end{aligned}
$$

These substitutions reproduce the fifth column, up to display rounding,
using exactly the same timing as the full MLE calculation. What has changed
is the restriction on the long-run level. Fixing that level makes the
estimation problem smaller and can improve numerical stability, provided
that the sample variance is a reasonable anchor for the period under study.
The comparison below examines how much fit is lost by imposing this restriction.

It also requires us to state how the estimated quantities are counted in
AIC and BIC. Our main calculation treats $V_L$ as fixed during optimization
and uses $k=2$ for the two fitted weights. Yet $V_L$ itself was obtained from
the training sample, so we also report a count of three that includes this
sample-based input. Showing both conventions allows us to separate the
empirical difference in fit from the choice of complexity penalty.

Before comparing the objectives, it is useful to place the two daily variance
paths alongside the same observed returns. The table below keeps the dates,
day numbers, and price levels used in the individual calculation tables,
and places the two estimates of $\hat v_i$ in adjacent columns. Their shared
initialization makes day 3 identical. From day 4 onward, the small differences
reflect the fitted coefficients, since both procedures receive exactly the
same return history. These columns report conditional variances; taking
their square roots would give conditional volatilities.

```{r compare-daily-garch}
daily_comparison <- compact_garch_calculation(full_daily) |>
  select(date, day, price, return, variance_full = variance) |>
  mutate(variance_targeting = compact_garch_calculation(vt_daily)$variance)

stopifnot(identical(full_daily$date, vt_daily$date))
daily_comparison |>
  kable(
    caption = "Daily GARCH(1,1) conditional variances: full maximum likelihood and variance targeting.",
    col.names = c("Date", "Day", "S&P 500", "\\(u_i\\)",
                  "\\(\\hat v_i\\) (Full MLE)", "\\(\\hat v_i\\) (Variance targeting)"),
    escape = FALSE, row.names = FALSE, align = c("l", rep("r", 5))
  ) |>
  kable_styling(full_width = FALSE, latex_options = "HOLD_position")
```

## Objective surface and estimator comparison

Having reduced the optimization to two free parameters, we can visualize
the search itself. Each point in the $(\alpha,\beta)$ plane determines an
intercept through variance targeting and hence a complete variance path and
likelihood value. The contours join pairs giving the same fit, while the
marked solution identifies the pair returned by `optim()`. We plot the
Gaussian log-likelihood so that the contour labels use the same scale as
the reported results. Converting from the proportional score preserves both
the contour shapes and the optimum's location; the selected axis limits let
us examine their behavior near that solution.

```{r}
#| label: fig-garch-vt-contour
#| fig-cap: "Variance targeting objective-function contours."
alpha_grid <- seq(0.22, 0.30, length.out = 100)
beta_grid <- seq(0.66, 0.76, length.out = 100)

objective_surface <- outer(alpha_grid, beta_grid, Vectorize(function(alpha, beta) {
  if (alpha < 0 || beta < 0 || (alpha + beta) >= 0.999) {
    return(NA_real_)
  }

  nll_garch_vt(c(alpha, beta), u = u, long_run_variance = long_run_variance)
}))

proportional_levels <- c(10700, 10750, 10800, 10820, 10830, 10835)
loglik_surface <-
  -objective_surface / 2 - n_vt / 2 * log(2 * pi)
contour_levels <-
  proportional_levels / 2 - n_vt / 2 * log(2 * pi)

contour_lines <- grDevices::contourLines(
  x = beta_grid,
  y = alpha_grid,
  z = t(loglik_surface),
  levels = contour_levels
)

contour_tbl <- bind_rows(lapply(seq_along(contour_lines), function(id) {
  tibble(
    contour_id = id,
    beta = contour_lines[[id]]$x,
    alpha = contour_lines[[id]]$y,
    log_likelihood = contour_lines[[id]]$level
  )
}))

label_candidates <- contour_tbl |>
  filter(
    beta >= 0.662,
    beta <= 0.756,
    alpha >= 0.225,
    alpha <= 0.285
  )

level_tbl <- tibble(
  log_likelihood = contour_levels,
  label_beta_target = c(0.715, 0.730, 0.743, 0.748, 0.752, 0.755)
)

contour_label_tbl <- label_candidates |>
  inner_join(level_tbl, by = "log_likelihood") |>
  group_by(log_likelihood) |>
  slice_min(abs(beta - label_beta_target), n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(label = scales::number(log_likelihood, accuracy = 0.1, big.mark = ""))

solution_tbl <- tibble(
  beta = beta_vt,
  alpha = alpha_vt
)

contour_tbl |>
  ggplot(aes(beta, alpha)) +
  geom_path(aes(group = contour_id), color = "#1B4E8A", linewidth = 0.5) +
  geom_label(
    data = contour_label_tbl,
    aes(label = label),
    color = "#2F2F2F",
    fill = "white",
    label.size = 0,
    size = 3,
    alpha = 0.9
  ) +
  geom_hline(yintercept = alpha_vt, color = "#B22222", linetype = 2) +
  geom_vline(xintercept = beta_vt, color = "#B22222", linetype = 2) +
  geom_point(
    data = solution_tbl,
    color = "#B22222",
    size = 2.8
  ) +
  coord_cartesian(xlim = c(0.66, 0.76), ylim = c(0.22, 0.30), expand = FALSE) +
  labs(
    x = expression(beta),
    y = expression(alpha),
    caption = "Contour labels show Gaussian log-likelihood values; higher values indicate better fit."
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

As the contour labels increase toward the red point, the fit improves and
the negative score minimized by `optim()` decreases. The shape also helps
us understand the search: a long, flat contour would indicate that some
changes in $\alpha$ can be offset by changes in $\beta$ with little loss
of fit. To assess the cost of anchoring the long-run variance, we now compare
this solution with the one obtained when $\omega$ is estimated freely.

```{r compare-garch}
comparison_tbl <- bind_rows(
  full_summary |> mutate(model = "Full MLE"),
  vt_summary |> mutate(model = "Variance targeting")
) |>
  pivot_wider(names_from = model, values_from = value) |>
  select(metric, `Full MLE`, `Variance targeting`)

comparison_tbl |>
  kable(
    caption = "Full MLE and variance targeting comparison.",
    digits = 6,
    col.names = c("Metric", "Full MLE", "Variance targeting"),
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")

optimization_audit_tbl <- tibble(
  model = c("Full MLE", "Variance targeting"),
  convergence_code = c(fit_full$convergence, fit_vt$convergence),
  proportional_criterion = c(-fit_full$value, -fit_vt$value),
  gaussian_logLik = c(loglik_full, loglik_vt)
)

optimization_audit_tbl |>
  kable(
    caption = "Optimization and likelihood audit.",
    digits = 6,
    col.names = c(
      "Model",
      "Convergence code",
      "Proportional criterion",
      "Gaussian logLik"
    ),
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")

aic_vt_k3 <- -2 * loglik_vt + 2 * 3
bic_vt_k3 <- -2 * loglik_vt + log(n_vt) * 3

parameter_count_tbl <- tibble(
  specification = c(
    "Full MLE",
    "Variance targeting: conditional count",
    "Variance targeting: count sample-based V_L"
  ),
  quantities_counted = c(
    "omega, alpha, beta",
    "alpha, beta",
    "V_L, alpha, beta"
  ),
  k = c(3, 2, 3),
  AIC = c(aic_full, aic_vt, aic_vt_k3),
  BIC = c(bic_full, bic_vt, bic_vt_k3)
)

parameter_count_tbl |>
  kable(
    caption = "AIC and BIC under two parameter counts for variance targeting.",
    digits = 6,
    col.names = c(
      "Specification",
      "Quantities counted",
      "k",
      "AIC",
      "BIC"
    ),
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

The two likelihoods are evaluated on the same `r n_full` observations,
so their difference measures fit on a common basis. Full MLE reaches
`r scales::number(loglik_full, accuracy = 0.001)`, compared with
`r scales::number(loglik_vt, accuracy = 0.001)` for variance targeting. The
gain from estimating the intercept freely is therefore
`r scales::number(loglik_full - loglik_vt, accuracy = 0.001)`. The ordering
also provides a numerical check: the variance-targeting parameter vector is
available to the unrestricted model, so the unrestricted maximum cannot be
lower. Both searches report convergence code zero, and the direct Gaussian
evaluation agrees with the conversion from the proportional score. Together,
these checks support reading the small difference as a feature of the fitted
models rather than a mismatch in calculations.

The similarity in fit carries over to the economic interpretation. Full MLE
has persistence `r scales::number(ab_full, accuracy = 0.001)` and a half-life
of `r scales::number(half_life_full, accuracy = 0.1)` trading days, compared
with `r scales::number(ab_vt, accuracy = 0.001)` and
`r scales::number(half_life_vt, accuracy = 0.1)` days under variance targeting.
Both therefore describe a persistent response to volatility shocks, with
the freely estimated model allowing that response to decay slightly faster.
Anchoring the long-run variance changes the fit modestly without changing
this broad reading of the dynamics.

The AIC and BIC comparison adds a penalty for the estimated quantities, which
is where the treatment of $V_L$ enters. Full MLE estimates three coefficients
directly and therefore has $k=3$. Variance targeting optimizes two weights
and obtains $\omega=V_L(1-\alpha-\beta)$, giving $k=2$ when the long-run
input is treated as fixed during that step. Including the fact that $V_L$
was itself estimated from these returns gives a count of three instead.
In either case, $k$ counts quantities entering the penalty; it is unrelated
to the lag orders denoted by $(1,1)$.

Under the conditional count of two, the smaller penalty gives variance
targeting the lower AIC and BIC. Counting $V_L$ puts both fits at $k=3$,
so Full MLE's slightly higher likelihood then gives it the lower criteria,
by `r scales::number(2 * (loglik_full - loglik_vt), accuracy = 0.001)` points
each. Changing the count leaves the fitted coefficients, likelihood,
variance paths, and forecasts untouched: the switch in preference comes
entirely from the penalty. Reporting both conventions makes that dependence
visible and puts the very small equal-count difference in perspective.

The information criteria thus give little reason to expect substantially
different descriptions of historical risk. To see what the estimates imply
in financial terms, we turn from the parameter comparison to the volatility
paths and forecasts they generate.

## Conditional volatility through time and forecasts {#sec-garch-forecast}

Taking the square root of each fitted variance translates the updating rule
into a daily measure of risk that can be read alongside the returns. We use
the variance-targeting fit for this illustration because its long-run level
is anchored explicitly to the sample variance, giving a natural reference
for the plot. The full MLE path can be reconstructed in the same way. Such
time-varying risk measures underpin the risk-management, scenario-analysis,
and derivatives applications discussed by @hull2022options and
@mcneil2015quantitative.

```{r build-volatility-series}
volatility_tbl <- returns |>
  mutate(
    variance_vt = v_vt,
    volatility_vt = sqrt(variance_vt),
    sample_volatility = sqrt(long_run_variance)
  )

volatility_tbl |>
  select(date, return, variance_vt, volatility_vt) |>
  slice(c(1:6, (n() - 1):n())) |>
  kable(
    caption = "Conditional variance and volatility from variance targeting.",
    col.names = c("Date", "\\(u_t\\)", "\\(\\hat v_t\\)", "\\(\\sqrt{\\hat v_t}\\)"),
    escape = FALSE,
    digits = 8,
    format.args = list(scientific = FALSE),
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

```{r}
#| label: fig-garch-volatility
#| fig-cap: "S&P 500 conditional daily volatility from GARCH(1,1) estimated by variance targeting."
volatility_tbl |>
  ggplot(aes(date, 100 * volatility_vt)) +
  geom_line(color = "#1B4E8A", linewidth = 0.6) +
  geom_hline(
    aes(yintercept = 100 * sample_volatility),
    color = "#B22222",
    linetype = 2
  ) +
  labs(
    x = "Date",
    y = "Daily volatility",
    caption = "Dashed line: sample daily volatility."
  ) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1, scale = 1)) +
  theme_minimal(base_size = 12)
```

The horizontal line averages over the full sample, whereas the fitted series
responds to the changing size of the observed returns. Its sharp increase
during the 2020 market stress episode shows how the recursion translates
successive large movements into a higher assessment of risk. During quieter
periods, smaller squared returns gradually bring that assessment down.

To move from filtering the historical data to forecasting, first recall
the GARCH updating rule at an ordinary date $t$:

$$
v_t=\omega+\alpha u_{t-1}^2+\beta v_{t-1}.
$$

The variance assigned to day $t$ uses the return and variance from day
$t-1$. At the final observation, $T$, advancing this same rule by one period
means replacing $t$ by $T+1$: the preceding return becomes $u_T$ and the
preceding variance becomes $v_T$. With fitted parameters, we already have
the return and the fitted variance needed for this update. Writing its
result as $\hat v_{T+1\mid T}$ makes the information date explicit: this
is tomorrow's variance forecast using information through today. Thus the
first forecast requires no assumption about a future realized shock:

$$
\hat{v}_{T+1\mid T}
=
\omega
+
\alpha u_T^2
+
\beta v_T.
$$

The next step introduces uncertainty because $u_{T+1}$ has not yet been
observed. Under the zero-mean model, its expected square equals its conditional
variance. Using that expectation in the recursion combines the two weights
into $\alpha+\beta$, and repeating the argument gives the forecast at
any horizon $h$:

$$
\hat{v}_{T+h\mid T}
=
\bar{v}
+
(\alpha+\beta)^{h-1}
\left(
\hat{v}_{T+1\mid T}-\bar{v}
\right),
\qquad h \ge 1.
$$

This expression begins at the one-step forecast and reduces its deviation
from the long-run variance, $\bar v$, by a factor of $\alpha+\beta$ for each
additional trading day. In the calculation below, the parameters and final
variance are replaced by their fitted values. Taking square roots then puts
the forecasts on the same volatility scale as the historical plot.

```{r garch-volatility-forecast}
last_return <- tail(u, 1)
last_variance <- tail(v_vt, 1)

v_next <- omega_vt + alpha_vt * last_return^2 + beta_vt * last_variance
long_run_vt <- omega_vt / (1 - alpha_vt - beta_vt)

horizon <- 20
forecast_variance <- long_run_vt +
  (alpha_vt + beta_vt)^(seq_len(horizon) - 1) * (v_next - long_run_vt)

forecast_tbl <- tibble(
  horizon = seq_len(horizon),
  variance_forecast = forecast_variance,
  volatility_forecast = sqrt(forecast_variance)
)

forecast_tbl |>
  kable(
    caption = "Twenty-day conditional variance and volatility forecasts from variance targeting.",
    col.names = c("Horizon", "\\(\\hat v_{T+h\\mid T}\\)", "\\(\\sqrt{\\hat v_{T+h\\mid T}}\\)"),
    escape = FALSE,
    digits = 8,
    format.args = list(scientific = FALSE),
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

```{r}
#| label: fig-garch-volatility-forecast
#| fig-cap: "Historical GARCH volatility and its twenty-trading-day forecast from variance targeting: full history and an enlarged view from January 2020. Shading shows a pointwise 95% simulated interval for future conditional volatility."
#| fig-width: 9
#| fig-height: 8
set.seed(20260923)
simulation_count <- 50000L
simulated_variance <- matrix(NA_real_, nrow = horizon, ncol = simulation_count)
simulated_variance[1, ] <- v_next
for (h in 2:horizon) {
  simulated_return <- sqrt(simulated_variance[h - 1, ]) * rnorm(simulation_count)
  simulated_variance[h, ] <- omega_vt + alpha_vt * simulated_return^2 +
    beta_vt * simulated_variance[h - 1, ]
}
forecast_interval <- forecast_tbl |>
  mutate(
    lower = apply(sqrt(simulated_variance), 1, quantile, probs = 0.025),
    upper = apply(sqrt(simulated_variance), 1, quantile, probs = 0.975),
    trading_day = nrow(returns) + horizon
  )

# Trading-day positions avoid inventing future prices or calendar observations.
history_plot <- volatility_tbl |>
  mutate(trading_day = row_number())
forecast_line <- bind_rows(
  tibble(trading_day = nrow(returns), volatility_forecast = sqrt(last_variance)),
  forecast_interval |> select(trading_day, volatility_forecast)
)
year_ticks <- history_plot |>
  group_by(year = format(date, "%Y")) |>
  slice_head(n = 1) |>
  ungroup()

zoom_history <- history_plot |> filter(date >= as.Date("2020-01-01"))
month_ticks <- zoom_history |>
  group_by(month = format(date, "%Y-%m")) |>
  slice_head(n = 1) |>
  ungroup()
panel_levels <- c("Full history", "From January 2020")
history_panels <- bind_rows(
  history_plot |> mutate(panel = panel_levels[1]),
  zoom_history |> mutate(panel = panel_levels[2])
) |>
  mutate(panel = factor(panel, levels = panel_levels))
interval_panels <- bind_rows(lapply(panel_levels, function(view) {
  forecast_interval |> mutate(panel = factor(view, levels = panel_levels))
}))
line_panels <- bind_rows(lapply(panel_levels, function(view) {
  forecast_line |> mutate(panel = factor(view, levels = panel_levels))
}))
forecast_end <- nrow(returns) + horizon

ggplot() +
  geom_line(
    data = history_panels,
    aes(trading_day, 100 * volatility_vt, color = "Historical"), linewidth = 0.6
  ) +
  geom_ribbon(
    data = interval_panels,
    aes(trading_day, ymin = 100 * lower, ymax = 100 * upper),
    fill = "#CC7A00", alpha = 0.20
  ) +
  geom_line(
    data = line_panels,
    aes(trading_day, 100 * volatility_forecast, color = "Forecast"), linewidth = 0.9
  ) +
  geom_vline(xintercept = nrow(returns), linetype = 3, color = "gray50") +
  geom_hline(
    yintercept = 100 * sqrt(long_run_vt),
    color = "#B22222", linetype = 2
  ) +
  scale_color_manual(values = c("Historical" = "#1B4E8A", "Forecast" = "#CC7A00")) +
  scale_x_continuous(
    breaks = function(limits) {
      ticks <- if (diff(limits) > 500) year_ticks$trading_day else month_ticks$trading_day
      c(ticks, forecast_end)
    },
    labels = function(ticks) {
      date_labels <- if (diff(range(ticks)) > 500) "%Y" else "%b"
      labels <- format(history_plot$date[match(ticks, history_plot$trading_day)], date_labels)
      labels[ticks == forecast_end] <- "T + 20"
      labels
    }
  ) +
  facet_wrap(~ panel, ncol = 1, scales = "free_x") +
  labs(
    x = "Trading-day sequence (T is the final training date)",
    y = "Daily volatility", color = NULL,
    caption = "Dashed horizontal line: long-run volatility. Dotted vertical line: forecast origin."
  ) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.1, scale = 1)) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

The upper panel places the forecast in the full historical record, while
the lower panel enlarges the period from January 2020 onward so that the
orange segment and its shaded interval can be examined. Both panels use
the same vertical scale and the same values; only the horizontal window
changes. Observations are spaced by trading day, with years or months marking
their position and $T+20$ indicating the final forecast horizon.

In the enlarged view, the smooth forecast is easier to distinguish from the
irregular historical path. Each historical update incorporates a newly
observed shock, so the fitted volatility rises and falls as market movements
change. At $T$, all later shocks are still unknown. Averaging over them leaves
the regular decay toward the long-run variance described by the forecast
equation. Because the last fitted variance is already relatively close to
that level, the projected volatility changes only gradually over these
20 days. Its smoothness reflects the information available when it is made.

The plotted line takes the square root of expected future variance, giving
the standard deviation of the future daily return conditional on information
at $T$ under the zero-mean model. That choice keeps the forecast on the same
percentage scale as historical volatility. It also identifies precisely what
is being plotted: taking an expectation and taking a square root in the
opposite order would generally give a different quantity.

The smooth line leaves open how much conditional volatility could vary as
future shocks arrive. To explore that question, we simulate 50,000 possible
paths using the fitted parameters and independent standard-normal shocks,
following the approach in @archForecasting. On each path, a shock is multiplied
by its conditional standard deviation to generate a return, which then feeds
into the next variance update. Collecting the resulting volatilities at a
given horizon lets us locate their 2.5th and 97.5th percentiles. Repeating
this across horizons produces the shaded pointwise 95% interval. The seed
fixes the simulation draws for reproducibility and plays no role in fitting
or selecting the model.

Every path begins with the same one-step variance because the final training
return and variance are already known. The interval consequently has zero
width at that step. Once simulated shocks enter subsequent updates, the
paths separate and their distribution can become asymmetric, which is why
we construct the band from percentiles. Since the coefficients are held
fixed, it represents uncertainty from future shocks under the fitted model
and excludes uncertainty about the coefficient estimates. Its 95% coverage
applies separately at each horizon; it neither describes an interval for
returns nor guarantees coverage of a whole volatility path. These distinctions
will also be important when we compare the forecast with what happens after
the training period.

## Standardized Residual Diagnostics {#sec-garch-diagnostics}

Before evaluating the forecasts, we can use the training sample to check
whether the fitted model has accounted for the dependence that motivated it.
The return plot showed periods of unusually large movements; if the variance
updates capture that behavior, returns measured relative to their fitted risk
should exhibit less clustering. This suggests dividing each observed return
by the conditional standard deviation assigned to it:

$$
z_t
=
\frac{u_t}{\sqrt{\hat{v}_t}}.
$$

Here $u_t$ is the residual from our zero-conditional-mean equation, and
division by $\sqrt{\hat v_t}$ removes its estimated time-varying scale.
A value of $z_t=2$ means that the return was two fitted standard deviations
above zero; $z_t=-2$ means two below. A movement during a quiet period can
therefore be judged on the same basis as one during a turbulent period.
This construction uses the return because the data contain no observed
"true conditional volatility" from which we could subtract the fitted
volatility. Under the correctly specified model with known parameters,
the standardized shocks would have conditional mean zero and variance one,
and the additional normality assumption would make them standard normal.

Standardization gives us two related ways to examine what the model has left
unexplained. If $z_t$ remains autocorrelated, earlier returns still contain
linear information about later ones, which can point to a weakness in the
zero-mean specification. If its square remains autocorrelated, large
standardized movements still tend to follow other large movements, suggesting
that the variance equation has left some clustering behind. Examining both
series separates these mean and variance concerns, which could easily be
confused by looking only at the original returns.

For each series, we use the Ljung-Box statistic to assess the first 20
autocorrelations jointly. A small p-value gives evidence against their being
zero, complementing the individual lag patterns in the plots. We apply the
same calculation on the same valid dates to both estimators, using
`Box.test()` with `fitdf = 0`. The reported p-values should be read as
approximate diagnostics: the usual chi-squared reference distribution does
not fully account for estimating a GARCH model, and formal residual inference
may require a model-specific adjustment or bootstrap. Even favorable results
would address only the dependence examined here, leaving normality,
independence, and forecasting performance to separate assessments.

```{r garch-diagnostics}
diag_tbl <- bind_rows(
  tibble(date = returns$date, return = u, variance = v_full, model = "Full MLE"),
  tibble(date = returns$date, return = u, variance = v_vt, model = "Variance targeting")
) |>
  filter(is.finite(variance), variance > 0) |>
  mutate(
    z = return / sqrt(variance), z_squared = z^2,
    model = factor(model, levels = c("Full MLE", "Variance targeting"))
  ) |>
  pivot_longer(c(z, z_squared), names_to = "series", values_to = "value")

diag_tests <- diag_tbl |>
  group_by(model, series) |>
  group_modify(function(.x, .y) {
    test <- stats::Box.test(.x$value, lag = 20, type = "Ljung-Box", fitdf = 0)
    tibble(statistic = unname(test$statistic), p_value = test$p.value)
  }) |>
  ungroup()

diag_tests |>
  mutate(series = ifelse(series == "z", "\\(z_t\\)", "\\(z_t^2\\)")) |>
  kable(
    caption = "Ljung-Box diagnostics at 20 lags for both GARCH estimators (approximate p-values).",
    col.names = c("Estimator", "Series", "Statistic", "p-value"), escape = FALSE,
    digits = 4,
    row.names = FALSE
  ) |>
  kable_styling(latex_options = "HOLD_position")
```

```{r}
#| label: fig-garch-residual-acf
#| fig-cap: "Autocorrelations of standardized residuals and their squares for both estimators. Dashed lines are approximate pointwise white-noise reference limits."
#| fig-height: 6
acf_tbl <- diag_tbl |>
  group_by(model, series) |>
  group_modify(function(.x, .y) {
    ac <- stats::acf(.x$value, plot = FALSE, lag.max = 20)
    tibble(lag = as.numeric(ac$lag[-1]), acf = as.numeric(ac$acf[-1]),
           limit = 1.96 / sqrt(nrow(.x)))
  }) |>
  ungroup() |>
  mutate(series = factor(series, levels = c("z", "z_squared"),
                         labels = c("Standardized returns", "Squared standardized returns")))

acf_tbl |>
  ggplot(aes(lag, acf)) +
  geom_col(width = 0.08, fill = "#1B4E8A") +
  geom_hline(data = distinct(acf_tbl, model, series, limit),
             aes(yintercept = limit), linetype = 2, color = "#B22222") +
  geom_hline(data = distinct(acf_tbl, model, series, limit),
             aes(yintercept = -limit), linetype = 2, color = "#B22222") +
  facet_grid(series ~ model) +
  scale_x_continuous(breaks = c(1, 5, 10, 15, 20)) +
  labs(x = "Lag", y = "ACF") +
  theme_minimal(base_size = 12) +
  theme(
    panel.spacing.x = grid::unit(1.2, "cm"),
    panel.spacing.y = grid::unit(0.6, "cm"),
    panel.border = element_rect(color = "grey65", fill = NA, linewidth = 0.5),
    strip.background = element_rect(fill = "grey95", color = NA)
  )
```

The columns place the two estimators side by side, while the rows distinguish
standardized returns from their squares. Their similar patterns agree with
the similar parameter estimates found earlier. Some individual spikes cross
the dashed pointwise reference limits, which can happen by chance when many
lags are examined. The joint Ljung-Box checks help put those crossings in
context: all four approximate p-values exceed 0.05, with the smallest equal
to `r sprintf("%.3f", min(diag_tests$p_value))`. At this 20-lag horizon,
neither estimator leaves clear evidence of the linear dependence being tested
in standardized returns or their squares.

These results support the fitted dynamics on the training sample, but we
still need to examine predictions on later observations. The dashed lines
here assess sampling variation in autocorrelations; the band in
@sec-garch-forecast describes possible future conditional volatilities.
Neither calculation, by itself, tells us how accurate a variance forecast
will be when confronted with new returns.

## Out-of-sample forecast evaluation {#sec-garch-evaluation}

So far, the observed returns have served both to estimate the parameters and
to assess the residuals produced by that fit. A forecast faces a further
test: it must describe observations that were unavailable when it was made.
We create that separation by stopping estimation at the existing final
training date, `r max(sp500$date)`, and evaluating forecasts against the
next 20 S&P 500 returns. The additional data enter only this evaluation;
all earlier estimates and calculations continue to use the original sample.

The practical question is what each estimator would have predicted before
each test return became available, and how those predictions compare with
the movements subsequently observed. We examine three forecasting policies
for each estimator. The first commits to all twenty forecasts at the end
of training. The second updates tomorrow's forecast each day using the
latest return while retaining the original coefficients. The third also
re-estimates the coefficients before issuing each daily forecast. Keeping
these policies distinct lets us separate the contribution of fresh return
information from the additional effect of learning new parameter values.

We obtain the next 20 closing observations from the same provider and
calculate the first test return relative to the final training close. This
retains the price change across the boundary while keeping its return
entirely in the test period. The first two policies hold both parameter
vectors and the variance-targeting input $V_L$ at their original training
values. The third uses an expanding estimation sample: a test return can
enter estimation only after its date has passed. Thus the return being
predicted remains excluded from the information used for that prediction,
even when earlier test returns have become available for re-estimation.

```{r garch-test-data}
training_end <- max(sp500$date)
test_prices <- tq_get(
  "^GSPC", from = as.character(training_end + 1),
  to = as.character(training_end + 60)
) |>
  select(date, close) |>
  rename(price = close) |>
  arrange(date) |>
  filter(date > training_end) |>
  slice_head(n = 20)

stopifnot(
  nrow(test_prices) == 20L, !anyDuplicated(test_prices$date),
  all(is.finite(test_prices$price)), all(test_prices$price > 0),
  all(diff(test_prices$date) > 0)
)
test_returns <- test_prices |>
  mutate(
    return = price / lag(price, default = tail(sp500$price, 1)) - 1,
    horizon = row_number(), squared_return = return^2
  )

tibble(
  sample = c("Training returns", "Test returns"),
  first_date = c(min(returns$date), min(test_returns$date)),
  last_date = c(max(returns$date), max(test_returns$date)),
  observations = c(nrow(returns), nrow(test_returns))
) |>
  kable(caption = "Chronological separation of training and test returns.",
        col.names = c("Sample", "First date", "Last date", "Returns"), row.names = FALSE) |>
  kable_styling(full_width = FALSE, latex_options = "HOLD_position")
```

The dates in the table show the resulting split. The test runs from
`r min(test_returns$date)` through `r max(test_returns$date)`, covering
20 trading observations rather than 20 calendar days. Its boundary follows
the last price actually included in training, so the download cutoff cannot
silently shift a return from one sample to the other.

We begin by asking how the arrival of these returns changes the model's
assessment of risk. Keep the 20 forecasts originally made at $T$ unchanged,
and alongside them calculate a second path that incorporates each test
return once it has been observed. Denoting this updated path by
$\tilde v_{T+h}$ distinguishes it from the fixed-origin forecast
$\hat v_{T+h\mid T}$. For either fitted parameter vector, the update is

$$
\begin{aligned}
\tilde v_{T+1}&=\hat v_{T+1\mid T},\\
\tilde v_{T+h}&=\hat\omega+\hat\alpha u_{T+h-1}^{2}
+\hat\beta\tilde v_{T+h-1},\qquad h=2,\ldots,20.
\end{aligned}
$$

For the first test day, both paths use the final training return and variance,
so their values coincide. For the second day, the updated path can use the
first test return, whereas the original forecast must still use its expectation
at $T$. The same distinction continues through the window. Although we
reconstruct the updated path after observing the data, every update uses
returns only through the preceding day. It is therefore also the sequence
of one-step forecasts that could have been issued as the test unfolded,
without changing the coefficients. Comparing it with the fixed-origin path
shows the effect of new information within the model; both paths remain
model-based estimates of an unobserved variance.

This timing also explains why the two lines can separate substantially.
For the second test day, subtracting the original forecast from the updated
variance gives

$$
\tilde v_{T+2}-\hat v_{T+2\mid T}
=\hat\alpha\left(u_{T+1}^2-\hat v_{T+1\mid T}\right).
$$

The original forecast used the expected squared return for day $T+1$.
Once that return is observed, the update replaces the expectation by its
realized square. A smaller squared return therefore lowers the next day's
updated variance relative to the original forecast; a larger one raises it.
On subsequent days, the model also carries forward the variance changes
caused by earlier surprises. A sequence of relatively quiet returns can
therefore pull the updated path well below the original forecast even
though the coefficients remain unchanged. Conversely, the fixed-origin
path keeps moving toward its fitted long-run level because it receives no
new information. This distinction applies separately to both estimators.

```{r garch-test-paths}
test_garch_paths <- function(omega, alpha, beta, last_variance, model) {
  h <- test_returns$horizon
  first_variance <- omega + alpha * tail(u, 1)^2 + beta * last_variance
  long_run <- omega / (1 - alpha - beta)
  fixed_origin <- long_run + (alpha + beta)^(h - 1) * (first_variance - long_run)
  filtered <- numeric(length(h))
  filtered[1] <- first_variance
  for (i in 2:length(h)) {
    filtered[i] <- omega + alpha * test_returns$return[i - 1]^2 + beta * filtered[i - 1]
  }
  test_returns |>
    mutate(model = model, fixed_origin = as.numeric(fixed_origin), filtered = filtered)
}

test_paths <- bind_rows(
  test_garch_paths(omega_full, alpha_full, beta_full, tail(v_full, 1), "Full MLE"),
  test_garch_paths(omega_vt, alpha_vt, beta_vt, tail(v_vt, 1), "Variance targeting")
) |>
  mutate(model = factor(model, levels = c("Full MLE", "Variance targeting")))

stopifnot(isTRUE(all.equal(
  test_paths$fixed_origin[test_paths$model == "Variance targeting"],
  as.numeric(forecast_variance), tolerance = 1e-12
)))
```

```{r}
#| label: fig-garch-test-paths
#| fig-cap: "Fixed-origin GARCH volatility forecasts and daily updated conditional volatilities during the test period. Both paths are shown as daily percentages, as in Section 3.9. Parameters remain fixed at their training estimates."
test_paths |>
  pivot_longer(c(fixed_origin, filtered), names_to = "path", values_to = "variance") |>
  mutate(path = factor(path, levels = c("fixed_origin", "filtered"),
                       labels = c("Forecast at T", "Updated with observed returns"))) |>
  ggplot(aes(date, 100 * sqrt(variance), color = path)) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ model) +
  scale_color_manual(values = c("#CC7A00", "#1B4E8A")) +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 0.01, scale = 1)) +
  labs(x = "Test date", y = "Daily volatility", color = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

The orange line in each panel preserves what could be forecast at $T$,
while the blue line shows how that estimator responds to the returns as
they arrive. For continuity with @sec-garch-forecast, both paths are shown
as daily volatility percentages: we take the square root of each variance
and multiply by 100. A value of 1% therefore has the same meaning in these
figures. The shorter test window permits a closer view of the movement,
so its axis limits differ from those of the full historical plot while
the units remain the same. This display transformation leaves the underlying
variance forecasts and all subsequent loss calculations unchanged.
Within each panel, the divergence between the lines comes entirely from
the information update; their similarity across panels reflects how close
the two estimated models are.

By the last test day, the updated daily volatility is approximately
`r sprintf("%.3f", 100 * sqrt(tail(test_paths$filtered[test_paths$model == "Full MLE"], 1)))`%
for Full MLE and
`r sprintf("%.3f", 100 * sqrt(tail(test_paths$filtered[test_paths$model == "Variance targeting"], 1)))`%
for variance targeting. The original forecasts for
that day remain at
`r sprintf("%.3f", 100 * sqrt(tail(test_paths$fixed_origin[test_paths$model == "Full MLE"], 1)))`%
and
`r sprintf("%.3f", 100 * sqrt(tail(test_paths$fixed_origin[test_paths$model == "Variance targeting"], 1)))`%.
The returns observed over the window thus lead both models to a lower
end-of-period assessment of risk. This illustrates their reaction to new
information, but evaluating forecast accuracy requires a separate observable
outcome. Using the filtered path itself as the outcome would merely compare
two outputs of the same fitted model.

The third policy adds parameter re-estimation to this daily information
update. Before predicting day $t$, we extend the estimation sample through
$t-1$, fit the model, and reconstruct its variance path with those newly
estimated coefficients. Denote the coefficients by
$\hat\omega_{t-1}$, $\hat\alpha_{t-1}$, and $\hat\beta_{t-1}$, where the
subscript identifies the last observation available for estimation. The
resulting one-day forecast is

$$
\hat v_{t\mid t-1}^{\mathrm{refit}}
=\hat\omega_{t-1}+\hat\alpha_{t-1}u_{t-1}^2
+\hat\beta_{t-1}\hat v_{t-1}^{\mathrm{refit}}.
$$

Here $\hat v_{t-1}^{\mathrm{refit}}$ is the final filtered variance obtained
by running the recursion over the expanded sample with the new coefficients.
Reconstructing this path keeps the variance state consistent with the model
just estimated. For variance targeting, we also recalculate $V_L$ from the
expanded sample and derive the new intercept from that value. Full MLE
continues to estimate all three coefficients jointly. Neither procedure
uses the return on day $t$ until after forecasting that day.

These policies answer different operational questions. The original
twenty-day forecast describes what can be planned at the initial date,
before any test returns arrive. Daily updating describes a risk assessment
that can be revised as the market evolves while keeping a previously
calibrated model. Daily re-estimation also allows the calibration to change,
at the cost of repeating the optimization and introducing estimation
variation. Comparing the first two policies measures the benefit of fresh
information in this window; comparing the last two isolates the additional
effect of re-estimating the model on an expanding sample.

```{r garch-daily-refits}
daily_refits <- list()
for (h in seq_len(nrow(test_returns))) {
  past <- c(u, head(test_returns$return, h - 1L))
  if (h == 1L) {
    pf <- unname(theta_full)
    pv <- c(unname(omega_vt), unname(alpha_vt), unname(beta_vt))
  } else {
    full_fit <- optim(c(4e-6, 0.2, 0.7), nll_garch, u = past,
      method = "L-BFGS-B", lower = c(1e-12, 0, 0),
      upper = c(Inf, 1 - 1e-6, 1 - 1e-6),
      control = list(parscale = c(1e-6, 0.1, 0.1), factr = 1e7, pgtol = 1e-10))
    target <- var(past)
    vt_fit <- optim(c(0.2, 0.7), nll_garch_vt, u = past,
      long_run_variance = target, method = "L-BFGS-B",
      lower = c(0, 0), upper = c(1 - 1e-6, 1 - 1e-6))
    # Retry an unsuccessful search using the previous fit and finer numerical steps.
    if (vt_fit$convergence != 0) {
      vt_fit <- optim(unname(pv[2:3]), nll_garch_vt, u = past,
        long_run_variance = target, method = "L-BFGS-B",
        lower = c(0, 0), upper = c(1 - 1e-6, 1 - 1e-6),
        control = list(parscale = c(0.1, 0.1), ndeps = c(1e-5, 1e-5), maxit = 2000))
    }
    stopifnot(full_fit$convergence == 0, vt_fit$convergence == 0)
    pf <- unname(full_fit$par)
    pv <- c(target * (1 - sum(vt_fit$par)), unname(vt_fit$par))
    stopifnot(nll_garch(pf, past) <= nll_garch(pv, past) + 1e-6)
  }
  for (model in c("Full MLE", "Variance targeting")) {
    p <- if (model == "Full MLE") pf else pv
    stopifnot(all(p >= 0), p[1] > 0, sum(p[2:3]) < 1)
    last_v <- tail(garch_variance(past, p[1], p[2], p[3]), 1)
    prediction <- p[1] + p[2] * tail(past, 1)^2 + p[3] * last_v
    daily_refits[[length(daily_refits) + 1L]] <- tibble(
      date = test_returns$date[h], model = model, forecast = prediction,
      squared_return = test_returns$squared_return[h],
      policy = "Daily re-estimation", training_n = length(past),
      omega = p[1], alpha = p[2], beta = p[3])
  }
}
daily_refits <- bind_rows(daily_refits)

forecast_comparison <- test_paths |>
  select(date, model, squared_return, fixed_origin, filtered) |>
  pivot_longer(c(fixed_origin, filtered), names_to = "policy", values_to = "forecast") |>
  mutate(policy = ifelse(policy == "fixed_origin", "Original 20-day forecast", "Daily update, fixed parameters")) |>
  bind_rows(daily_refits |> select(date, model, squared_return, policy, forecast)) |>
  mutate(policy = factor(policy, levels = c("Original 20-day forecast",
    "Daily update, fixed parameters", "Daily re-estimation")))

stopifnot(all(is.finite(forecast_comparison$forecast)), all(forecast_comparison$forecast > 0))
first_day_check <- forecast_comparison |>
  filter(date == min(date)) |>
  group_by(model) |>
  summarise(gap = max(forecast) - min(forecast), .groups = "drop")
stopifnot(all(first_day_check$gap < 1e-12))
```

The expression `head(test_returns$return, h - 1L)` is the chronological
safeguard in this calculation: before test day $h$, only the preceding
$h-1$ test returns can be appended to training. On the first test day,
all three policies consequently give the same forecast for each estimator.
Later re-estimations retain the original initialization rule and check
convergence, parameter restrictions, and likelihood ordering. If the initial
variance-targeting search fails to converge, the code retries from the
previous day's estimate with finer numerical steps; an unsuccessful retry
stops the calculation. These numerical checks concern estimation, without
using test losses to select parameter values.

To evaluate all three policies, we turn to the squared return $u_{T+h}^2$. Under the
zero-conditional-mean assumption, its conditional expectation equals the
day's latent variance, making it a useful observable proxy. Any one squared
return can nevertheless lie far above or below that variance. A quiet
realization is possible on a risky day, just as a large movement can occur
on a day assigned relatively low risk. We must consequently choose a loss
function suited to a noisy proxy and interpret averages over only 20
observations cautiously. The role of that choice is examined in
@patton2011volatility.

For this second comparison, the next figure uses variance units because
the observable outcome is a squared return. Both the forecast variance and
the squared return are multiplied by 10,000 to express them in squared
percentage points. For example, a daily volatility of 1% corresponds to a
variance of 1 squared percentage point; a volatility of 2% corresponds to
4 squared percentage points. Each panel shows the variance forecast issued
under each policy and plots the squared
return actually observed on each test date. All six panels use the same
observed returns, so differences in forecast performance arise from the
forecast lines. The variation of the points around those lines also shows
why a single squared return provides such an imprecise assessment of that
day's underlying risk.

```{r}
#| label: fig-garch-test-proxy
#| fig-cap: "Three forecasting policies compared with observed squared returns. Columns distinguish the policies; rows distinguish full MLE and variance targeting. All panels use the same test dates and variance scale."
#| fig-width: 12
#| fig-height: 7
forecast_comparison |>
  ggplot(aes(date)) +
  geom_line(aes(y = 10000 * forecast, color = "Variance forecast"), linewidth = 0.8) +
  geom_line(aes(y = 10000 * squared_return, color = "Observed squared return"), linewidth = 0.35) +
  geom_point(aes(y = 10000 * squared_return, color = "Observed squared return"), size = 1.5) +
  facet_grid(model ~ policy) +
  scale_color_manual(values = c("Variance forecast" = "#CC7A00", "Observed squared return" = "#1B4E8A")) +
  scale_x_date(date_breaks = "2 weeks", date_labels = "%b %d") +
  labs(x = "Test date", y = "Daily variance (squared percentage points)", color = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.spacing = grid::unit(0.7, "cm"),
        strip.text.x = element_text(size = 10),
        panel.border = element_rect(color = "grey65", fill = NA, linewidth = 0.5))
```

The squared returns vary much more abruptly than the forecast paths because
each point records one realized daily movement, whereas each forecast
describes an expected squared movement using the information available
when that forecast was issued.
The updated variances underlying the preceding volatility figure smooth
these movements because the recursion combines the previous squared return
with the previous variance. They also respond with a one-day lag: a return
observed today enters tomorrow's conditional variance. These differences
in timing and meaning account for the different shapes of the figures.
The loss calculations below aggregate the discrepancies in this second
comparison on the same test dates. Each forecast is aligned with the return
it was issued to predict; the observed series is connected by a thin line
to make its chronological sequence easier to follow.

For each estimator and policy, let $f_h$ be its variance forecast for test
day $T+h$, issued using the information allowed by that policy, and let
$q_h=u_{T+h}^2$ be the subsequently observed proxy. We score their discrepancy
using two average losses, QLIKE and mean squared error:

$$
\begin{aligned}
\mathrm{QLIKE}&=\frac{1}{20}\sum_{h=1}^{20}
  \left[\log(f_h)+\frac{q_h}{f_h}\right],\\
\mathrm{MSE}&=\frac{1}{20}\sum_{h=1}^{20}(q_h-f_h)^2.
\end{aligned}
$$

The form of QLIKE should look familiar from the likelihood calculation.
Assigning a large variance raises $\log(f_h)$, while assigning too little
variance to a large observed squared return raises $q_h/f_h$. Their sum
balances the two costs, now judging forecasts already issued rather than
choosing parameters to fit the sample. Terms depending only on the proxy
can be omitted because they would add the same amount to each model's loss
on the same dates. This version also remains defined when a return is zero.
Its value can be negative, so performance is read by comparing which loss
is lower, regardless of the sign.

MSE supplies a more direct distance measure by squaring the gap between the
forecast variance and the squared return. It gives substantial weight to
large discrepancies and has units equal to the fourth power of decimal
returns. To make the small values readable, the table multiplies MSE by
$10^8$, equivalent to calculating it from variances in squared percentage
points. Under a conditionally unbiased proxy and the required moment
conditions, both criteria have useful expected-ranking properties
[@patton2011volatility]. Their use makes the comparison appropriate for this
target, while the noise in a short test sample remains.

```{r garch-test-losses}
evaluation_summary <- test_paths |>
  group_by(model) |>
  summarise(
    observations = n(),
    qlike = mean(log(fixed_origin) + squared_return / fixed_origin),
    mse = mean((squared_return - fixed_origin)^2),
    mean_forecast = mean(fixed_origin),
    mean_proxy = mean(squared_return),
    .groups = "drop"
  )

policy_scores <- forecast_comparison |>
  group_by(model, policy) |>
  summarise(observations = n(),
    qlike = mean(log(forecast) + squared_return / forecast),
    mse = mean((squared_return - forecast)^2), .groups = "drop")

policy_scores |>
  transmute(model, policy, qlike, mse_scaled = 1e8 * mse) |>
  kable(caption = "Forecast losses across the three policies: the same twenty test dates for each estimator and policy.",
        col.names = c("Estimator", "Forecasting policy", "QLIKE", "\\(\\mathrm{MSE}\\times10^8\\)"),
        digits = c(0, 0, 6, 6), escape = FALSE, row.names = FALSE) |>
  kable_styling(full_width = FALSE, latex_options = "HOLD_position")

qlike_best <- as.character(evaluation_summary$model[which.min(evaluation_summary$qlike)])
mse_best <- as.character(evaluation_summary$model[which.min(evaluation_summary$mse)])
loss_reading <- if (qlike_best == mse_best) {
  paste(qlike_best, "has the lower loss under both QLIKE and MSE")
} else {
  paste(qlike_best, "has the lower QLIKE, while", mse_best, "has the lower MSE")
}

update_gains <- policy_scores |>
  select(model, policy, mse) |>
  pivot_wider(names_from = policy, values_from = mse) |>
  mutate(reduction = 100 * (1 - `Daily update, fixed parameters` / `Original 20-day forecast`))
```

For the original twenty-day forecasts, `r loss_reading`. The QLIKE values are
`r sprintf("%.6f", evaluation_summary$qlike[1])` for Full MLE and
`r sprintf("%.6f", evaluation_summary$qlike[2])` for variance targeting;
the corresponding scaled MSE values are
`r sprintf("%.6f", 1e8 * evaluation_summary$mse[1])` and
`r sprintf("%.6f", 1e8 * evaluation_summary$mse[2])`.
To understand this ordering, compare the risk forecasts with the size of the
observed movements. The average squared return is
`r sprintf("%.3f", 1e4 * evaluation_summary$mean_proxy[1])` in
squared-percentage-point units, compared with average variance forecasts of
`r sprintf("%.3f", 1e4 * evaluation_summary$mean_forecast[1])` and
`r sprintf("%.3f", 1e4 * evaluation_summary$mean_forecast[2])`.
Both models therefore assign more variance on average than the observed
proxy in this window. Full MLE projects slightly less variance, and its
forecasts yield smaller discrepancies under both loss functions. That is
the empirical result for these dates; it gives no basis for claiming a
statistically significant or lasting advantage over variance targeting.

The larger improvement in the table comes from updating the forecasts
each day. Relative to the original twenty-day path, daily updating with
fixed parameters reduces MSE by
`r sprintf("%.1f", update_gains$reduction[update_gains$model == "Full MLE"])`%
for full MLE and
`r sprintf("%.1f", update_gains$reduction[update_gains$model == "Variance targeting"])`%
for variance targeting. QLIKE also falls for both estimators. In the middle
column of the figure, the forecasts move downward as the quieter returns
arrive, bringing them closer to the subsequent squared returns on average.
Some large discrepancies remain because the forecast describes conditional
variance, while each squared return records just one draw from that day's
return distribution. The improvement is measured over all twenty dates;
it does not imply that every individual forecast becomes more accurate.

Daily re-estimation changes the picture very little beyond that update.
The last two columns are almost indistinguishable, and their losses are
very close. Full MLE improves slightly under both criteria when re-estimated;
variance targeting has a marginally lower MSE but a marginally higher
QLIKE. There is consequently little evidence here of an additional practical
benefit from re-estimating every day. The original estimation sample already
contains `r length(u)` returns. Before the final test forecast, it has grown
by only 19 returns, or `r sprintf("%.1f", 100 * 19 / length(u))`% of its
original size. The twentieth test return becomes available only after the
last forecast has been issued. These few additions change the fitted
coefficients little in this sample, whereas the daily variance recursion
responds directly to each latest squared return. This explains why most
of the adaptation comes from updating the variance state. A shorter
estimation window or a different market episode could yield a more
substantial effect from re-estimation.

The comparison therefore separates two decisions: whether forecasts can
incorporate new observations, and whether the model is recalibrated when
they do. Its interpretation must respect the information advantage of the
daily policies. The original policy makes forecasts at horizons one through
twenty from one date; the daily policies make twenty successive one-step
forecasts. Lower daily-policy losses describe the benefit of that forecasting
practice in this window, rather than a contest between procedures supplied
with identical information. Within each policy, full MLE and variance
targeting use the same information dates and can be compared directly.

A broader assessment would repeat this design over additional test windows
and market conditions. Twenty daily outcomes provide limited evidence,
especially with a noisy variance proxy. The favorable residual checks,
the small in-sample likelihood difference, and the test losses answer
related questions about remaining dependence, fit, and predictive accuracy.
Together they give a fuller assessment than either visual agreement or a
single loss ranking alone.

## Common implementation mistakes

Several implementation errors can produce plausible-looking results while
breaking the link between the model and the forecast. An especially easy
one is to use $u_t^2$ in the calculation of $v_t$. That would let the model
use the very return whose variance it is supposed to predict. Working through
the first rows of the daily calculation table helps expose the mistake:
the update must use $u_{t-1}^2$ and $v_{t-1}$, and the initialized value must
appear before the first recursive update. The price-row numbers also need
to remain distinct from the indices of the return vector after its first
missing observation has been removed.

A similarly unobtrusive error is to mix returns expressed as percentages
with variances calculated from decimal returns. Multiplying a return by
100 multiplies its variance by 10,000, so the mismatch changes the likelihood
as well as the apparent magnitude of risk. Keeping decimal units throughout
estimation and converting only for presentation makes it easier to check
that the squared returns and conditional variances are on the same scale.

Even with correct timing and units, the optimizer's output needs scrutiny.
The parameter vector must satisfy $\omega>0$, $\alpha\ge0$, $\beta\ge0$,
and $\alpha+\beta<1$ for the positive, finite-long-run-variance model used
here. The coordinate bounds do not enforce the final joint restriction,
which is why it is also checked inside the objective. Numerical convergence
then needs to be interpreted alongside the attained likelihood. Trying
several feasible starting vectors should give similar estimates in a stable
problem, and scaling the parameters helps avoid search difficulties caused
by their different magnitudes. An additional check is available in this
application: the unrestricted likelihood should be at least as high as the
likelihood at any admissible variance-targeting solution. A successful
convergence code alone would not reveal a violation of that ordering.

Once a fit has passed these checks, a comparison can still be misleading
if the reported statistic uses the wrong likelihood scale or observation
set. The proportional score is sufficient for estimating parameters, but
`gaussian_loglik()` supplies the complete value required by AIC and BIC.
Both models must be evaluated on the same valid dates, with the treatment
of the sample-based long-run variance made explicit in the parameter count.
Otherwise an apparent preference can arise from the reporting convention
rather than a substantive difference between models.

The interpretation needs the same care as the calculation. Information
criteria describe a penalized fit and should be considered together with
parameter plausibility and residual behavior. A claim about forecasting
performance additionally requires later observations that were excluded
from fitting. In that assessment, keeping the filtered GARCH path separate
from the observed squared-return proxy prevents a comparison between model
outputs from being mistaken for validation against data. These are the
connections that allow the numerical results to answer the financial
question posed at the start of the chapter.

## Conclusion and extensions

We began with the observation that the size of daily S&P 500 movements changes
over time. GARCH gives that observation a quantitative form by allowing the
latest squared return to revise a variance estimate that also retains part
of its preceding value. Following the recursion through the daily tables
showed how this rule connects the data to the likelihood. Maximizing that
likelihood then provided the coefficients needed to reconstruct historical
volatility and project risk beyond the end of the sample.

The two estimators illustrate how a restriction can simplify this process
without greatly changing its empirical description. Full MLE chooses
$(\omega,\alpha,\beta)$ jointly, while variance targeting estimates the
two weights and derives the intercept from a sample-based long-run variance.
Both optimizations converge, and the freely estimated model attains the
slightly higher Gaussian likelihood that its larger feasible set permits.
Their persistent variance dynamics and similar standardized-residual
diagnostics support broadly comparable readings of historical risk. The
AIC/BIC preference depends on how the sample-based input is counted:
variance targeting is favored under the conditional count $k=2$, while
counting $V_L$ gives $k=3$ and a narrow preference for Full MLE. Given the
small difference, this comparison provides little evidence of a substantial
gap in fit.

Forecasting carries the same recursion forward with a change in information.
Historical variance responds to observed shocks; a forecast averages over
shocks still to come, which explains its smoother approach to the long-run
level. The simulated band around the variance-targeting forecast shows how
future conditional volatility can nevertheless evolve along quite different
paths with the fitted coefficients held fixed. The target throughout is
the variability of future returns, so a useful risk forecast need not imply
an ability to predict their direction.

The subsequent observations, from `r min(test_returns$date)` through
`r max(test_returns$date)`, let us distinguish that original prediction from
the updates made possible by new returns. Both models revised their
end-of-window risk assessment downward as the test unfolded. Comparing
three forecasting policies showed that daily updating with fixed coefficients
substantially reduced both loss criteria relative to the original twenty-day
forecast. Re-estimating each day added little in this window, where the
expanding sample gained at most 19 returns before the final forecast,
alongside `r length(u)` original training returns. Full MLE and variance
targeting produced very similar paths under each policy. These findings
illustrate the value of new information for daily risk assessment while
keeping the distinction between one-step and multi-step forecasts explicit.
They concern one short window with a noisy variance proxy; a general claim
of forecasting superiority would require evaluation over additional windows
and market conditions. Read together, the
estimation results, residual checks, and test losses show both what this
simple model can explain and the limits of the evidence used to assess it.

Further developments can be motivated by the assumptions that remain.
An asymmetric response to gains and losses would lead toward EGARCH or
GJR-GARCH, heavier conditional tails toward Student-t innovations, and
joint risk across assets toward multivariate models such as DCC-GARCH.
These extensions are discussed in @francq2019garch, @tsay2010analysis, and
@mcneil2015quantitative. Each changes part of the model, but the route from
an economic question to an explicit variance equation, an estimable
criterion, and an honest forecast evaluation remains the basis for
understanding and using the result.
