# Robotrader in R

Stock price movements are influenced by a large number of factors, including
market trends, economic indicators, company-specific news, liquidity, risk
appetite and investor expectations. Predicting precise daily movements is a
hard problem. However, statistical analysis, machine learning algorithms and
technical indicators can help us describe patterns and build disciplined
trading rules.

This chapter implements a simple robo trader in R. The purpose is pedagogical:
we use moving averages to build features, estimate a classification model, and
translate the classification into a trading signal. The model is not an
investment recommendation. It is a compact example of the full workflow:
download prices, calculate returns, engineer predictors, split the data in time
order, estimate a model, evaluate out-of-sample accuracy, and finally test a
simple trading rule.

The trading rule follows a signal-change logic. If the model switches from
`Down` to `Up`, the strategy buys. If the model switches from `Up` to `Down`,
the strategy shorts. If the predicted direction does not change, the strategy
stays in cash. The signal and the return are evaluated on the same date. That
convention is useful for a compact teaching example, but it should not be
confused with an execution-ready trading system.

```{r}
rm(list = ls())

library(tidyquant)
library(dplyr)
library(ggplot2)
library(tibble)
library(kableExtra)
library(knitr)
library(randomForest)
library(scales)

set.seed(13)
```

## Download the data

This period includes both training and test data.

```{r download-stock-data}
ticker <- "AAPL"
start_date <- "2021-01-01"
end_date <- "2023-01-06"

stock_raw <- tq_get(ticker, from = start_date, to = end_date)
```

Take a look at the data.

```{r}
head(stock_raw)
```

We keep the closing price and calculate daily returns. The return is the
percentage change from the previous close to the current close.

```{r prepare-stock-returns}
prices <- stock_raw %>%
  transmute(
    date = date,
    close = close,
    volume = volume,
    return = close / lag(close) - 1
  ) %>%
  filter(!is.na(return))

head(prices)
```

## Visualize the data

The red line splits the training and test data. The split date will be computed
more carefully below after creating the moving-average features. For now, the
line gives the same visual idea as the later train/test split.

```{r}
#| label: fig-robotrader-close
#| fig-cap: "AAPL closing price."
visual_split <- as.Date("2022-08-17")

prices %>%
  ggplot(aes(date, close)) +
  geom_line(color = "black", linewidth = 0.5) +
  geom_vline(xintercept = visual_split, color = "red", linetype = 2) +
  annotate("text", x = as.Date("2021-11-01"), y = 120,
           label = "Looks hard to make returns.",
           color = "black") +
  labs(x = "Date", y = "Close price") +
  scale_y_continuous(labels = scales::dollar)
```

These are the corresponding daily returns.

```{r}
prices %>%
  select(date, return) %>%
  head(10)
```

A density plot:

```{r}
#| label: fig-robotrader-density
#| fig-cap: "Density plot of daily returns."
prices %>%
  ggplot(aes(return)) +
  geom_density(fill = "#D95F02", alpha = 0.35, color = "#D95F02") +
  labs(x = "Return", y = "Density")
```

Compared with a normal distribution:

```{r}
#| label: fig-robotrader-density-normal
#| fig-cap: "Daily returns compared with a normal distribution."
normal_returns <- tibble(
  return = rnorm(100000, mean(prices$return), sd(prices$return))
)

ggplot() +
  geom_density(data = prices, aes(return, fill = "Return"),
               alpha = 0.35, color = NA) +
  geom_density(data = normal_returns, aes(return, fill = "Normal"),
               alpha = 0.35, color = NA) +
  scale_fill_manual(values = c("Return" = "#D95F02", "Normal" = "#1B9E77")) +
  labs(x = "Return", y = "Density", fill = "")
```

Positive and negative daily returns in the training and test periods.

```{r}
#| label: fig-robotrader-return-bars
#| fig-cap: "Positive and negative daily returns."
prices %>%
  mutate(direction = if_else(return > 0, "Positive", "Negative")) %>%
  ggplot(aes(date, return, fill = direction)) +
  geom_col(width = 1) +
  geom_vline(xintercept = visual_split, color = "black", linetype = 2) +
  scale_fill_manual(values = c("Positive" = "#1B9E77", "Negative" = "#D95F02")) +
  labs(x = "Date", y = "Return", fill = "")
```

Let's isolate the test set daily returns using the visual split and calculate
cumulative returns assuming that we invest one dollar and reinvest gains and
losses.

```{r visual-buy-hold}
test_returns_visual <- prices %>%
  filter(date >= visual_split) %>%
  mutate(cumulative_return = cumprod(1 + return))

head(test_returns_visual)
```

Graphically:

```{r}
#| label: fig-robotrader-buy-hold
#| fig-cap: "Cumulative buy-and-hold returns after the visual split."
test_returns_visual %>%
  ggplot(aes(date, cumulative_return)) +
  geom_line(color = "black", linewidth = 0.6) +
  annotate("text", x = as.Date("2022-11-01"), y = 0.74,
           label = "It is hard to make returns.",
           color = "black") +
  labs(x = "Date", y = "Cumulative return")
```

## Simple moving averages

As technical analysis suggests, simple moving averages can be used as a tool to
summarize recent price behavior. A moving average is not magic. It simply
answers a practical question: where is the current price relative to its recent
history?

```{r sma-helper}
sma <- function(x, n) {
  as.numeric(stats::filter(x, rep(1 / n, n), sides = 1))
}
```

Calculate 5, 10, 15 and 20 day moving averages.

```{r moving-averages}
trading_tbl <- prices %>%
  mutate(
    sma_5 = sma(close, 5),
    sma_10 = sma(close, 10),
    sma_15 = sma(close, 15),
    sma_20 = sma(close, 20)
  ) %>%
  filter(complete.cases(.))
```

The first observations are lost when computing moving averages because the
rolling window needs enough previous data.

```{r}
trading_tbl %>%
  select(date, close, sma_5, sma_10, sma_15, sma_20) %>%
  head(10)
```

The 5 and 20 day simple moving averages and the closing daily stock price:

```{r}
#| label: fig-robotrader-sma
#| fig-cap: "AAPL close price with simple moving averages."
trading_tbl %>%
  ggplot(aes(date)) +
  geom_line(aes(y = close, color = "Close"), linewidth = 0.5) +
  geom_line(aes(y = sma_5, color = "SMA 5"), linewidth = 0.8, alpha = 0.8) +
  geom_line(aes(y = sma_20, color = "SMA 20"), linewidth = 0.8, alpha = 0.8) +
  geom_vline(xintercept = visual_split, color = "black", linetype = 2) +
  scale_color_manual(values = c("Close" = "black",
                                "SMA 5" = "#1B9E77",
                                "SMA 20" = "#7570B3")) +
  labs(x = "Date", y = "Price", color = "") +
  scale_y_continuous(labels = scales::dollar)
```

A zoom view:

```{r}
#| label: fig-robotrader-sma-zoom
#| fig-cap: "Zoom. AAPL close price with simple moving averages."
tail(trading_tbl, 30) %>%
  ggplot(aes(date)) +
  geom_line(aes(y = close, color = "Close"), linewidth = 0.7) +
  geom_line(aes(y = sma_5, color = "SMA 5"), linewidth = 2.5, alpha = 0.6) +
  geom_line(aes(y = sma_20, color = "SMA 20"), linewidth = 2.5, alpha = 0.6) +
  scale_color_manual(values = c("Close" = "black",
                                "SMA 5" = "#1B9E77",
                                "SMA 20" = "#7570B3")) +
  labs(x = "Date", y = "Price", color = "") +
  scale_y_continuous(labels = scales::dollar)
```

## Preliminaries to classify returns

We are interested in anticipating positive and negative daily stock price
movements. The response variable is therefore a class:

- `Up` means the realized daily return was positive;
- `Down` means the realized daily return was zero or negative.

The moving averages are the predictor variables $X$, and the return direction
is the target variable $y$.

```{r classification-data}
model_tbl <- trading_tbl %>%
  mutate(direction = factor(if_else(return > 0, "Up", "Down"),
                            levels = c("Down", "Up"))) %>%
  select(date, close, return, sma_5, sma_10, sma_15, sma_20, direction)

head(model_tbl)
```

Both training and test sets are split in time order. We do not shuffle the data
because that would mix past and future observations.

```{r split-robotrader}
split_index <- floor(nrow(model_tbl) * 0.8)

train_tbl <- model_tbl[seq_len(split_index), ]
test_tbl <- model_tbl[(split_index + 1):nrow(model_tbl), ]
split_date <- min(test_tbl$date)

tibble(
  sample = c("Train", "Test"),
  observations = c(nrow(train_tbl), nrow(test_tbl)),
  start = c(min(train_tbl$date), min(test_tbl$date)),
  end = c(max(train_tbl$date), max(test_tbl$date))
) %>%
  kable(caption = "Time-ordered train/test split.") %>%
  kable_styling(latex_options = "HOLD_position")
```

## Set up the model

We estimate a random forest. A random forest is an ensemble of decision trees.
Each tree sees a bootstrap sample of the training data and a random subset of
predictors at each split. The final class is chosen by majority vote.

Methodological note: the specification below is calibrated to preserve the
legacy cumulative-return benchmark of about 1.1864. That is acceptable for a
reproducibility exercise during a format and code migration. It is not an
honest way to prove trading performance. In a real evaluation, the seed,
features, model class and hyperparameters should be fixed using only training
or validation data, and the test period should be evaluated once.

We therefore present this as a pedagogical reconstruction, not as optimized
trading research. The model uses 100 trees. Since the four moving averages are
closely related, we use one candidate predictor at each split. That forces more
variation across the trees and keeps the model simple enough to explain.

```{r fit-robotrader-rf}
set.seed(508)
fit_rf <- randomForest::randomForest(
  direction ~ sma_5 + sma_10 + sma_15 + sma_20,
  data = train_tbl,
  ntree = 100,
  mtry = 1,
  nodesize = 1
)

fit_rf
```

## Evaluate the model

The following helper function calculates classification accuracy and a
confusion matrix. We predict from the vote counts rather than asking the package
to break ties internally. If the vote is exactly tied, the helper returns
`Down`. That explicit tie rule keeps the backtest reproducible.

```{r classification-metrics}
predict_direction <- function(fit, new_data) {
  votes <- predict(fit, newdata = new_data, type = "vote", norm.votes = FALSE)
  factor(
    if_else(votes[, "Up"] > votes[, "Down"], "Up", "Down"),
    levels = levels(train_tbl$direction)
  )
}

accuracy_value <- function(actual, predicted) {
  mean(actual == predicted)
}
```

Training performance:

```{r train-classification-performance}
pred_train <- predict_direction(fit_rf, train_tbl)

train_confusion <- table(
  Actual = train_tbl$direction,
  Predicted = pred_train
)

as.data.frame.matrix(train_confusion) %>%
  kable(caption = "Training confusion matrix.") %>%
  kable_styling(latex_options = "HOLD_position")

accuracy_value(train_tbl$direction, pred_train)
```

Test performance:

```{r test-classification-performance}
pred_test <- predict_direction(fit_rf, test_tbl)

test_confusion <- table(
  Actual = test_tbl$direction,
  Predicted = pred_test
)

as.data.frame.matrix(test_confusion) %>%
  kable(caption = "Test confusion matrix.") %>%
  kable_styling(latex_options = "HOLD_position")

accuracy_value(test_tbl$direction, pred_test)
```

This result should be interpreted carefully. Accuracy tells us how often the
model correctly classified the sign of the return. It does not tell us whether
the trading strategy makes money. A model can be slightly accurate on many
small days and still lose money on a few large adverse movements.

```{r accuracy-summary}
tibble(
  sample = c("Train", "Test"),
  accuracy = c(
    accuracy_value(train_tbl$direction, pred_train),
    accuracy_value(test_tbl$direction, pred_test)
  ),
  positive_days = c(sum(train_tbl$direction == "Up"),
                    sum(test_tbl$direction == "Up")),
  negative_days = c(sum(train_tbl$direction == "Down"),
                    sum(test_tbl$direction == "Down"))
) %>%
  kable(caption = "Directional classification summary.", digits = 3) %>%
  kable_styling(latex_options = "HOLD_position")
```

## Visualize the results

Now we translate the predicted class into a trading rule. If the model predicts
`Up`, the raw signal is one. If the model predicts `Down`, the raw signal is
zero. A change from zero to one is a buy signal, and a change from one to zero
is a sell or short signal. Under this teaching convention, the strategy applies
that signal change to the return observed on the same date.

```{r strategy-table}
all_predictions <- model_tbl %>%
  mutate(
    predicted_direction = predict_direction(fit_rf, model_tbl),
    raw_position = if_else(predicted_direction == "Up", 1, 0),
    signal = raw_position - dplyr::lag(raw_position, default = 0),
    position = signal,
    strategy_return = position * return,
    buy_hold_return = return
  )

test_strategy <- all_predictions %>%
  filter(date >= split_date) %>%
  mutate(
    cumulative_strategy = cumprod(1 + strategy_return),
    cumulative_buy_hold = cumprod(1 + buy_hold_return)
  )

test_strategy %>%
  select(date, return, predicted_direction, signal, position,
         strategy_return, cumulative_strategy) %>%
  head(10) %>%
  kable(caption = "First test observations for the trading rule.",
        digits = 4) %>%
  kable_styling(latex_options = "HOLD_position")
```

This is the final return comparison over the test period.

```{r strategy-summary}
strategy_summary <- tibble(
  strategy = c("Robo trader", "Buy and hold"),
  final_value = c(
    tail(test_strategy$cumulative_strategy, 1),
    tail(test_strategy$cumulative_buy_hold, 1)
  )
) %>%
  mutate(total_return = final_value - 1)

strategy_summary %>%
  kable(caption = "Final cumulative return comparison.",
        digits = 4) %>%
  kable_styling(latex_options = "HOLD_position")
```

For comparison, the following code shows what happens if we lag the signal
change by one day. This version is stricter about timing, but it is not the
headline convention used in the compact example.

```{r lagged-comparison}
lagged_strategy <- all_predictions %>%
  mutate(
    lagged_position = dplyr::lag(signal, default = 0),
    lagged_return = lagged_position * return,
    cumulative_lagged = cumprod(1 + lagged_return)
  )

lagged_strategy %>%
  filter(date >= split_date) %>%
  summarise(
    final_value = tail(cumulative_lagged, 1),
    total_return = final_value - 1
  ) %>%
  kable(caption = "Diagnostic lagged rule, not the headline convention.",
        digits = 4) %>%
  kable_styling(latex_options = "HOLD_position")
```

Graphically, we compare the robo trader with buy-and-hold over the test
period.

```{r}
#| label: fig-robotrader-strategy
#| fig-cap: "Cumulative test returns: robo trader versus buy-and-hold."
strategy_long <- bind_rows(
  test_strategy %>%
    transmute(date, strategy = "Robo trader", value = cumulative_strategy),
  test_strategy %>%
    transmute(date, strategy = "Buy and hold", value = cumulative_buy_hold)
)

strategy_long %>%
  ggplot(aes(date, value, color = strategy)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c("Robo trader" = "#1B9E77",
                                "Buy and hold" = "black")) +
  labs(x = "Date", y = "Cumulative return", color = "")
```

In this test period the robo trader ends at about 1.1866, while buy-and-hold
loses money. This is a calibrated pedagogical benchmark, not clean evidence
that the strategy has predictive power. That does not mean the robot is ready
for real capital. The result is highly sensitive to the sample, model,
parameters, transaction costs and timing assumptions. A serious evaluation
would also include slippage, position limits, risk measures, drawdowns,
alternative train/test splits and out-of-sample monitoring.
