---
execute:
  freeze: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.pos = "H",
  out.extra = ""
)
```

```{r book-edition, echo=FALSE, results='asis'}
source("R/book-edition.R")
write_book_edition_script()
```

```{r contact-links, echo=FALSE, results='asis'}
xfun::file_string("contact.html")
```

```{r publication-dates, echo=FALSE, results='asis'}
write_book_dates()
```

## Preface {.unnumbered}

::: {.home-panel}
This material relies on @hull2020machine, @hyndman2021forecasting, and on Matt Dancho Business Science IO freely available codes in R that explain how to implement forecasting and machine learning workflows. Some mathematical background is skipped to emphasize the data analysis, model logic, discussion, graphical approach and R coding [@Rref]. As in the philosophy of @knuth1984literate, the objective of this document is to explain to human beings what we want a computer to do as literate programming. This is a work in progress and it is under revision.
:::

## What's new in this edition {.unnumbered}

- **Book format:** The material is now organized as a Quarto book with chapter navigation, search, and a cleaner reading flow.
- **Separated chapters:** The fpp3, transparent machine learning, and R robotrader material now live in individual chapters instead of one long rendered document.
- **Reader tools:** Code folding, copy buttons, local section navigation, reading progress, and back-to-top controls are available across the book.
